---
title: "Function guide"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Function guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r mock-data, include=FALSE}
library(encodeUtils)

json_response <- function(body) {
  httr2::response(
    200,
    headers = "Content-Type: application/json",
    body = charToRaw(body)
  )
}

rows <- function(x, ...) {
  as.data.frame(encode_results(x, ...), stringsAsFactors = FALSE)
}

show_columns <- function(x, columns) {
  x <- as.data.frame(x, stringsAsFactors = FALSE)
  x[, intersect(columns, names(x)), drop = FALSE]
}

show_experiments <- function(x) {
  show_columns(
    rows(x),
    c(
      "accession", "assay_title", "organism", "biosample_term_name",
      "biosample_type", "biosample_summary", "lab", "project",
      "file_count", "date_released", "status"
    )
  )
}

show_files <- function(x) {
  show_columns(
    rows(x),
    c(
      "experiment_accession", "file_accession", "assay_title",
      "biosample_term_name", "file_format", "output_type",
      "file_size_pretty", "md5sum", "status"
    )
  )
}

show_plan <- function(x) {
  show_columns(
    encode_results(x),
    c(
      "experiment_accession", "file_accession", "file_format",
      "output_type", "file_size_pretty", "local_path"
    )
  )
}

show_explain <- function(x) {
  show_columns(
    encode_explain_selection(x),
    c("file_accession", "experiment_accession", "decision", "reason")
  )
}

download_marker <- paste0(intToUtf8(64), intToUtf8(64), "download")

experiment_search_json <- paste0(
  '{"@graph":[{"accession":"ENCSR000AAA",',
  '"@id":"/experiments/ENCSR000AAA/",',
  '"@type":["Experiment"],',
  '"assay_title":"total RNA-seq","status":"released",',
  '"organism":{"scientific_name":"Homo sapiens"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"Homo sapiens heart tissue",',
  '"lab":{"title":"Example lab"},"award":{"project":"ENCODE"},',
  '"date_released":"2024-01-01","files":[1]}],"total":1}'
)

experiment_object_json <- paste0(
  '{"@type":["Experiment","Item"],"accession":"ENCSR000AAA",',
  '"@id":"/experiments/ENCSR000AAA/",',
  '"assay_title":"total RNA-seq","status":"released",',
  '"organism":{"scientific_name":"Homo sapiens"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"Homo sapiens heart tissue",',
  '"lab":{"title":"Example lab"},"award":{"project":"ENCODE"},',
  '"date_released":"2024-01-01"}'
)

file_search_json <- paste0(
  '{"@graph":[{"accession":"ENCFF000AAA",',
  '"@id":"/files/ENCFF000AAA/",',
  '"dataset":"/experiments/ENCSR000AAA/",',
  '"assay_title":"total RNA-seq",',
  '"simple_biosample_summary":"heart tissue",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"txt","output_type":"metadata",',
  '"file_size":3,"md5sum":"900150983cd24fb0d6963f7d28e17f72",',
  '"preferred_default":true,',
  '"href":"/files/ENCFF000AAA/', download_marker, '/ENCFF000AAA.txt",',
  '"status":"released"}],"total":1}'
)

matrix_json <- paste0(
  '{"total":3,"filters":[],',
  '"matrix":{"x":{"group_by":"assay_title",',
  '"assay_title":{"buckets":[{"key":"total RNA-seq","doc_count":2},',
  '{"key":"ChIP-seq","doc_count":1}]}},',
  '"y":{"group_by":["biosample_ontology.classification",',
  '"biosample_ontology.term_name"],',
  '"biosample_ontology.classification":{"buckets":[{"key":"tissue",',
  '"doc_count":3,"biosample_ontology.term_name":{"buckets":[{"key":"heart",',
  '"doc_count":3,"assay_title":{"buckets":[{"key":"total RNA-seq",',
  '"doc_count":2},{"key":"ChIP-seq","doc_count":1}]}}]}}]}}}}'
)
```

# Search records

`encode_search()` returns a result object that prints as a compact table.
`encode_results()` extracts the table.

```{r search}
res <- httr2::with_mocked_responses(
  function(req) json_response(experiment_search_json),
  encode_search(
    type = "Experiment",
    filters = list(assay_title = "total RNA-seq"),
    limit = 3,
    quiet = TRUE
  )
)
```

```{r search-table, echo=FALSE}
knitr::kable(show_experiments(res), row.names = FALSE)
```

# Check availability

`encode_matrix()` returns ENCODE Matrix counts by assay and biosample.

```{r matrix}
mat <- httr2::with_mocked_responses(
  function(req) json_response(matrix_json),
  encode_matrix(quiet = TRUE)
)
```

```{r matrix-tables, echo=FALSE}
knitr::kable(rows(mat), row.names = FALSE)
knitr::kable(rows(mat, component = "assays"), row.names = FALSE)
knitr::kable(rows(mat, component = "biosamples"), row.names = FALSE)
```

# Retrieve one record

`encode_get()` retrieves one ENCODE record by accession, path, or URL.

```{r get}
experiment <- httr2::with_mocked_responses(
  function(req) json_response(experiment_object_json),
  encode_get("ENCSR000AAA", quiet = TRUE)
)
```

```{r get-table, echo=FALSE}
knitr::kable(show_experiments(experiment), row.names = FALSE)
```

# Build a metadata table

`encode_report()` returns selected fields from matching records.

```{r report}
report <- httr2::with_mocked_responses(
  function(req) json_response(experiment_search_json),
  encode_report(
    fields = c("accession", "assay_title", "status"),
    limit = 1,
    quiet = TRUE
  )
)
```

```{r report-table, echo=FALSE}
knitr::kable(rows(report), row.names = FALSE)
```

# List files

`encode_list_files()` lists file metadata for experiments. It does not download
file contents.

```{r files}
files <- httr2::with_mocked_responses(
  function(req) json_response(file_search_json),
  encode_list_files(
    "ENCSR000AAA",
    file_format = "txt",
    quiet = TRUE
  )
)
```

```{r files-table, echo=FALSE}
knitr::kable(show_files(files), row.names = FALSE)
```

# Select files

`encode_select_files()` keeps selected rows and records why other rows were not
kept.

```{r select-files}
selected <- encode_select_files(
  files,
  preset = "metadata",
  prefer_default = TRUE,
  explain = FALSE
)
```

```{r select-files-table, echo=FALSE}
knitr::kable(show_files(selected), row.names = FALSE)
knitr::kable(show_explain(selected), row.names = FALSE)
```

# Preview and dry-run downloads

`encode_preview_download()` checks destination paths and file sizes.
`encode_download(..., dry_run = TRUE)` returns planned download rows without
transferring files.

```{r download}
plan <- encode_preview_download(
  selected,
  directory = tempdir(),
  quiet = TRUE
)

dry_run <- encode_download(
  selected,
  directory = tempdir(),
  dry_run = TRUE,
  quiet = TRUE
)
```

```{r download-tables, echo=FALSE}
knitr::kable(show_plan(plan), row.names = FALSE)
knitr::kable(show_plan(dry_run), row.names = FALSE)
```

# Record provenance

`encode_manifest()` records selected files, download rows, queries, and citation
metadata. `encode_cite()` returns dataset and file attribution.

```{r manifest}
manifest <- encode_manifest(
  selected,
  include_session = FALSE,
  path = file.path(tempdir(), "encode-manifest.json")
)
```

```{r manifest-tables, echo=FALSE}
knitr::kable(
  data.frame(section = names(manifest)),
  row.names = FALSE
)
knitr::kable(
  show_columns(
    encode_cite(selected, enrich = FALSE),
    c("accession", "type", "url", "dataset", "file_accession")
  ),
  row.names = FALSE
)
```

# Read local files

`encode_read()` reads supported files that are already on disk.

```{r read}
txt_path <- tempfile(fileext = ".txt")
writeLines("abc", txt_path)
encode_read(txt_path)

bam_path <- tempfile(fileext = ".bam")
writeBin(charToRaw("placeholder"), bam_path)
encode_read(bam_path)
```
