---
title: "Function guide"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Function guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(encodeUtils)
```

# Core functions

The main functions are:

- `encode_search()` to query ENCODE metadata;
- `encode_results()` to extract the main data frame from result objects;
- `encode_list_files()` to list files attached to experiment accessions;
- `encode_select_files()` to narrow file tables;
- `encode_download()` to plan or perform file downloads;
- `encode_read()` to read supported local files;
- `encode_manifest()` to record provenance.

The evaluated examples use small local records so the vignette remains stable
during package checks.

# Extract results

`encode_results()` returns the main table from an `encodeUtils` result object.
It also accepts plain data frames.

```{r results}
experiments <- data.frame(
  accession = c("ENCSR000AAA", "ENCSR000AAB"),
  assay_title = c("total RNA-seq", "ATAC-seq"),
  biosample_term_name = c("heart", "heart"),
  stringsAsFactors = FALSE
)

encode_results(experiments)
```

# Select files

`encode_select_files()` can use explicit filters or presets for common ENCODE
outputs.

```{r select}
files <- data.frame(
  experiment_accession = c("ENCSR000AAA", "ENCSR000AAA", "ENCSR000AAB"),
  file_accession = c("ENCFF000AAA", "ENCFF000AAB", "ENCFF000AAC"),
  file_format = c("tsv", "bed", "bigWig"),
  output_type = c(
    "gene quantifications",
    "IDR thresholded peaks",
    "signal p-value"
  ),
  assembly = c("mm10", "mm10", "mm10"),
  status = c("released", "released", "released"),
  file_size = c(120, 300, 5000000),
  md5sum = NA_character_,
  href = c(
    "/files/ENCFF000AAA/@@download/ENCFF000AAA.tsv",
    "/files/ENCFF000AAB/@@download/ENCFF000AAB.bed",
    "/files/ENCFF000AAC/@@download/ENCFF000AAC.bigWig"
  ),
  stringsAsFactors = FALSE
)

rna_selected <- encode_select_files(files, preset = "rnaseq_gene_quant")
peak_selected <- encode_select_files(files, preset = "atacseq_peaks")

encode_results(rna_selected)
encode_results(peak_selected)
```

# Plan downloads

Use `dry_run = TRUE` before transferring any file. The dry run reports planned
paths, sizes, and status without downloading contents.

```{r download-plan}
planned <- encode_download(
  rna_selected,
  directory = tempdir(),
  dry_run = TRUE,
  quiet = TRUE
)

encode_results(planned)[, c("file_accession", "download_status", "local_path")]
```

# Read Files

`encode_read()` can read small local tabular files directly. Gene
quantification tables are normalized to common identifier and expression columns
by default; set `simplify_quant = FALSE` to keep the raw downloaded columns.
BED-like interval files return `GRanges` when `GenomicRanges` and `IRanges` are
available, or a plain data frame when called with `as = "data.frame"`.

```{r read}
quant_path <- tempfile(fileext = ".tsv")
writeLines(
  c(
    "gene_id\tTPM\tFPKM\traw_counts",
    "ENSMUSG00000000001\t4.5\t2.1\t11",
    "ENSMUSG00000000028\t8.2\t4.4\t27"
  ),
  quant_path
)

downloaded <- encode_results(planned)
downloaded$local_path <- quant_path
downloaded$download_status <- "downloaded"
downloaded$organism <- "Mus musculus"

loaded <- encode_read(downloaded, values = c("raw_counts", "TPM"))
loaded$metadata
loaded$raw_counts

encode_read(quant_path, simplify_quant = FALSE)
```

# Manifest

`encode_manifest()` returns a list and can also write JSON when `path` is
provided.

```{r manifest}
manifest <- encode_manifest(downloaded, include_session = FALSE)
names(manifest)
```

# Live search and file listing

The live ENCODE examples are not evaluated during package checks. They are shown
as accession-aware patterns for interactive use.

```{r live-search, eval=FALSE}
experiments <- encode_search(
  type = "Experiment",
  organism = "mouse",
  assay = "histone chip-seq",
  organ = "heart",
  exclude_controls = TRUE,
  status = "released",
  limit = 5
)

files <- encode_list_files(
  experiments,
  file_format = "bed",
  assembly = "mm10",
  allow_many = TRUE
)

selected <- encode_select_files(
  files,
  preset = "chipseq_idr_peaks",
  assembly = "mm10"
)

encode_download(selected, directory = tempdir(), dry_run = TRUE)
```

# Session information

```{r session-info}
sessionInfo()
```
