---
title: "encodeUtils"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{1. encodeUtils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

```{r mock-data, include=FALSE}
library(encodeUtils)

json_response <- function(body) {
    httr2::response(
        200,
        headers = "Content-Type: application/json",
        body = charToRaw(body)
    )
}

experiment_search_json <- paste0(
    '{"@graph":[{"accession":"ENCSR000AAA",',
    '"@id":"/experiments/ENCSR000AAA/",',
    '"assay_title":"total RNA-seq","status":"released",',
    '"biosample_summary":"Homo sapiens heart tissue"}],"total":1}'
)

count_json <- paste0(
    '{"@graph":[],',
    '"total":128,',
    '"filters":[{"field":"type","term":"Experiment"},',
    '{"field":"assay_title","term":"total RNA-seq"}]}'
)

experiment_object_json <- paste0(
    '{"@type":["Experiment","Item"],"accession":"ENCSR000AAA",',
    '"@id":"/experiments/ENCSR000AAA/",',
    '"assay_title":"total RNA-seq","status":"released",',
    '"biosample_summary":"Homo sapiens heart tissue"}'
)

file_search_json <- paste0(
    '{"@graph":[{"accession":"ENCFF000AAA",',
    '"@id":"/files/ENCFF000AAA/",',
    '"dataset":"/experiments/ENCSR000AAA/",',
    '"assay_title":"total RNA-seq",',
    '"simple_biosample_summary":"heart tissue",',
    '"file_format":"txt","output_type":"metadata",',
    '"file_size":3,"md5sum":"900150983cd24fb0d6963f7d28e17f72",',
    '"preferred_default":true,',
    '"href":"/files/ENCFF000AAA/@@download/ENCFF000AAA.txt",',
    '"status":"released"}],"total":1}'
)

matrix_json <- paste0(
    '{"total":3,"filters":[],',
    '"matrix":{"x":{"group_by":"assay_title",',
    '"assay_title":{"buckets":[{"key":"total RNA-seq","doc_count":2},',
    '{"key":"ChIP-seq","doc_count":1}]}},',
    '"y":{"group_by":["biosample_ontology.classification",',
    '"biosample_ontology.term_name"],',
    '"biosample_ontology.classification":{"buckets":[{"key":"tissue",',
    '"doc_count":3,"biosample_ontology.term_name":{"buckets":[{"key":"heart",',
    '"doc_count":3,"assay_title":{"buckets":[{"key":"total RNA-seq",',
    '"doc_count":2},{"key":"ChIP-seq","doc_count":1}]}}]}}]}}}}'
)
```

# Overview

`encodeUtils` is a small R interface for ENCODE Portal metadata and selected
files. The package is organized around a deliberate workflow:

1. Search or summarize ENCODE metadata.
2. Inspect selected experiments or files.
3. List file metadata before downloading anything.
4. Summarize, select, and explain file decisions.
5. Preview destination paths, size lower bounds, and required overrides.
6. Download only selected files with size and checksum guardrails.
7. Record manifests and citation/provenance tables.
8. Read supported local files when that is safe.

For broad exploratory queries, `encode_count()` is available as an optional
safety preflight when you only need the live match count.

The package is read-only with respect to ENCODE. It does not implement
submission workflows.

# Search

This example searches a tiny page of released total RNA-seq experiments. The
printed object is concise, and `encode_results()` extracts the compact table for
downstream R work.

```{r search}
res <- httr2::with_mocked_responses(
    function(req) json_response(experiment_search_json),
    encode_search(
        filters = list(assay_title = "total RNA-seq"),
        limit = 3,
        quiet = TRUE
    )
)

encode_results(res)
```

# Optional Count Preflight

Most workflows can start with `encode_search()` because it already reports the
returned rows and live total. Use `encode_count()` for broad filters when you
want to check query size without fetching metadata rows.

```{r count}
count <- httr2::with_mocked_responses(
    function(req) json_response(count_json),
    encode_count(
        filters = list(assay_title = "total RNA-seq"),
        quiet = TRUE
    )
)

count
```

# Matrix Discovery

`encode_matrix()` is a quick way to ask what ENCODE has before narrowing a
search. It returns a long assay-by-biosample table plus assay and biosample
summaries.

```{r matrix}
mat <- httr2::with_mocked_responses(
    function(req) json_response(matrix_json),
    encode_matrix(quiet = TRUE)
)

encode_results(mat, component = "assays")
encode_results(mat, component = "biosamples")
```

# Inspect One Object

```{r get}
experiment <- httr2::with_mocked_responses(
    function(req) json_response(experiment_object_json),
    encode_get("ENCSR000AAA", quiet = TRUE)
)
encode_results(experiment)
```

# List Files Before Downloading

`encode_list_files()` is metadata only. It reports file accessions, formats,
sizes, MD5 checksums, replicate information, and download links.

```{r files}
files <- httr2::with_mocked_responses(
    function(req) json_response(file_search_json),
    encode_list_files(
        "ENCSR000AAA",
        file_format = "txt",
        quiet = TRUE
    )
)

encode_results(files)[, c("file_accession", "dataset_accession", "file_format", "file_size_pretty", "md5sum")]
```

# Summarize And Select Files

`encode_summary()` gives a compact overview before download. `encode_select_files()`
keeps both the selected files and an exclusion log so file decisions are
traceable.

```{r select-files}
encode_summary(files)

encode_select_files(preset = "metadata")

selected <- encode_select_files(
    files,
    preset = "metadata",
    prefer_default = TRUE,
    explain = FALSE
)

selected$files
selected$excluded
encode_explain_selection(selected)
```

# Plan A Download

Use `encode_preview_download()` before transfer to inspect destination paths,
known total size, unknown-size files, MD5 availability, and any required
overrides. `dry_run = TRUE` returns the same planned file paths in a table.

```{r download}
encode_preview_download(selected, directory = tempdir())

encode_download(
    selected,
    directory = tempdir(),
    dry_run = TRUE
)
```

# Manifest

`encode_manifest()` records package, query, selection, and file provenance in a
plain R list. Use `encode_write_manifest()` when you want a JSON copy alongside
downloaded files.

```{r manifest}
manifest <- encode_manifest(selected, include_session = FALSE)
names(manifest)
```

# Attribution

`encode_cite()` creates a provenance table for the specific ENCODE datasets and
files used downstream. Use `enrich = "auto"` to enrich small file tables with
parent experiment metadata, `enrich = TRUE` to force enrichment, or
`enrich = FALSE` to avoid extra web requests.

```{r cite}
encode_cite(selected, enrich = FALSE)

httr2::with_mocked_responses(
    function(req) json_response(experiment_search_json),
    encode_cite(selected, enrich = "auto", quiet = TRUE)
)
```

# Common Selection Presets

The preset names are aliases for explicit file-format and output-type
preferences. They do not download or read files.

```{r presets}
example_files <- data.frame(
    file_accession = c(
        "ENCFF000RNA", "ENCFF000TPM", "ENCFF000PKS",
        "ENCFF000SIG", "ENCFF000ATC", "ENCFF000FQ"
    ),
    experiment_accession = "ENCSR000AAA",
    file_format = c("tsv", "tsv", "bed", "bigWig", "bed", "fastq"),
    output_type = c(
        "gene counts", "gene TPMs", "optimal IDR thresholded peaks",
        "fold change over control", "IDR thresholded peaks", "reads"
    ),
    status = "released",
    href = "https://example.org/file.txt"
)

encode_select_files(example_files, preset = "rna_gene_counts", explain = FALSE)$files$file_accession
encode_select_files(example_files, preset = "rna_gene_tpm", explain = FALSE)$files$file_accession
encode_select_files(example_files, preset = "chipseq_idr_peaks", explain = FALSE)$files$file_accession
encode_select_files(example_files, preset = "chipseq_signal_bigwig", explain = FALSE)$files$file_accession
encode_select_files(example_files, preset = "atacseq_peaks", explain = FALSE)$files$file_accession
encode_select_files(example_files, preset = "raw_fastq", explain = FALSE)$files$file_accession
```

# Safe Local Reads

`encode_read()` reads small local text and JSON files. For formats that are easy
to misuse, such as FASTQ or full indexed signal files, it returns a lightweight
object explaining why it did not read the file by default.

```{r read}
fastq_path <- tempfile(fileext = ".fastq")
writeBin(charToRaw("placeholder"), fastq_path)
encode_read(fastq_path)
```

# Session Information

```{r session-info}
utils::sessionInfo()
```
