---
title: "Function guide"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Function guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r library}
library(encodeUtils)
```

# Core workflow

The core workflow is:

```{r core-workflow}
experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart total RNA-seq",
  status = "released",
  limit = 10
)

files <- encode_list_files(
  experiments,
  file_format = "tsv",
  output_type = "gene quantifications",
  assembly = "mm10"
)

file_ids <- c("ENCFF090VKE", "ENCFF260OJQ")

checked <- encode_download(
  files,
  file_accession = file_ids,
  directory = "data/encode/rna-seq",
  dry_run = TRUE
)

downloaded <- encode_download(
  files,
  file_accession = file_ids,
  directory = "data/encode/rna-seq"
)

loaded <- encode_read(downloaded)
manifest <- encode_manifest(downloaded, include_session = FALSE)
```

# Search

`encode_search()` finds ENCODE records and returns a compact result object.

```{r search}
experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart ChIP-seq",
  status = "released",
  limit = 10
)

encode_results(experiments)
```

# One record

`encode_get()` retrieves one ENCODE record by accession, portal path, or URL.

```{r get}
experiment <- encode_get("ENCSR284QGB")
encode_results(experiment)
```

# Availability counts

`encode_matrix()` returns ENCODE availability counts by assay and biosample.
Use it when you want a high-level view before writing a narrower search.

```{r matrix}
mat <- encode_matrix(
  filters = list("control_type!=" = "*", perturbed = "false")
)

encode_results(mat)
encode_results(mat, component = "assays")
encode_results(mat, component = "biosamples")
```

# Files

`encode_list_files()` lists downloadable files attached to experiments. It does
not download file contents.

```{r files}
files <- encode_list_files(
  experiments,
  file_format = "bed",
  assembly = "mm10"
)

encode_results(files)
```

# Selection

`encode_select_files()` narrows a file table. Use file accessions when you know
the exact files, or a preset when selecting a standard processed output.

```{r select}
selected <- encode_select_files(
  files,
  preset = "chipseq_idr_peaks",
  assembly = "mm10"
)

encode_results(selected)
encode_explain_selection(selected)
```

# Download and read

Use `dry_run = TRUE` first. Then run the same call without `dry_run` to transfer
files. `encode_read()` reads supported downloaded files without typing paths.

```{r download-read}
checked <- encode_download(
  selected,
  directory = "data/encode/chip-seq",
  dry_run = TRUE
)

downloaded <- encode_download(
  selected,
  directory = "data/encode/chip-seq"
)

loaded <- encode_read(downloaded)
loaded$metadata
```

# Manifest

`encode_manifest()` writes a reproducibility record for the query and files used
in an analysis. It includes ENCODE attribution metadata by default.

```{r manifest}
manifest <- encode_manifest(
  downloaded,
  include_session = FALSE,
  path = "data/encode/chip-seq/encode-manifest.json"
)

names(manifest)
```
