---
title: "Function guide"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Function guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r mock-data, include=FALSE}
library(encodeUtils)

json_response <- function(body) {
  httr2::response(
    200,
    headers = "Content-Type: application/json",
    body = charToRaw(body)
  )
}

download_marker <- paste0(intToUtf8(64), intToUtf8(64), "download")

experiment_search_json <- paste0(
  '{"@graph":[{"accession":"ENCSR000AAA",',
  '"@id":"/experiments/ENCSR000AAA/",',
  '"@type":["Experiment"],',
  '"assay_title":"total RNA-seq","status":"released",',
  '"organism":{"scientific_name":"Homo sapiens"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"Homo sapiens heart tissue",',
  '"lab":{"title":"Example lab"},"award":{"project":"ENCODE"},',
  '"date_released":"2024-01-01","files":[1]}],"total":1}'
)

experiment_object_json <- paste0(
  '{"@type":["Experiment","Item"],"accession":"ENCSR000AAA",',
  '"@id":"/experiments/ENCSR000AAA/",',
  '"assay_title":"total RNA-seq","status":"released",',
  '"organism":{"scientific_name":"Homo sapiens"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"Homo sapiens heart tissue",',
  '"lab":{"title":"Example lab"},"award":{"project":"ENCODE"},',
  '"date_released":"2024-01-01"}'
)

file_search_json <- paste0(
  '{"@graph":[{"accession":"ENCFF000AAA",',
  '"@id":"/files/ENCFF000AAA/",',
  '"dataset":"/experiments/ENCSR000AAA/",',
  '"assay_title":"total RNA-seq",',
  '"simple_biosample_summary":"heart tissue",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"txt","output_type":"metadata",',
  '"file_size":3,"md5sum":"900150983cd24fb0d6963f7d28e17f72",',
  '"preferred_default":true,',
  '"href":"/files/ENCFF000AAA/', download_marker, '/ENCFF000AAA.txt",',
  '"status":"released"}],"total":1}'
)

matrix_json <- paste0(
  '{"total":3,"filters":[],',
  '"matrix":{"x":{"group_by":"assay_title",',
  '"assay_title":{"buckets":[{"key":"total RNA-seq","doc_count":2},',
  '{"key":"ChIP-seq","doc_count":1}]}},',
  '"y":{"group_by":["biosample_ontology.classification",',
  '"biosample_ontology.term_name"],',
  '"biosample_ontology.classification":{"buckets":[{"key":"tissue",',
  '"doc_count":3,"biosample_ontology.term_name":{"buckets":[{"key":"heart",',
  '"doc_count":3,"assay_title":{"buckets":[{"key":"total RNA-seq",',
  '"doc_count":2},{"key":"ChIP-seq","doc_count":1}]}}]}}]}}}}'
)
```

# Search records

`encode_search()` returns a result object that prints as a compact table.
`encode_results()` extracts the table.

```{r search}
res <- httr2::with_mocked_responses(
  function(req) json_response(experiment_search_json),
  encode_search(
    type = "Experiment",
    filters = list(assay_title = "total RNA-seq"),
    limit = 3,
    quiet = TRUE
  )
)

res
encode_results(res)
```

# Check availability

`encode_matrix()` returns ENCODE Matrix counts by assay and biosample.

```{r matrix}
mat <- httr2::with_mocked_responses(
  function(req) json_response(matrix_json),
  encode_matrix(quiet = TRUE)
)

encode_results(mat)
encode_results(mat, component = "assays")
encode_results(mat, component = "biosamples")
```

# Retrieve one record

`encode_get()` retrieves one ENCODE record by accession, path, or URL.

```{r get}
experiment <- httr2::with_mocked_responses(
  function(req) json_response(experiment_object_json),
  encode_get("ENCSR000AAA", quiet = TRUE)
)

experiment
encode_results(experiment)
```

# Build a metadata table

`encode_report()` returns selected fields from matching records.

```{r report}
report <- httr2::with_mocked_responses(
  function(req) json_response(experiment_search_json),
  encode_report(
    fields = c("accession", "assay_title", "status"),
    limit = 1,
    quiet = TRUE
  )
)

encode_results(report)
```

# List files

`encode_list_files()` lists file metadata for experiments. It does not download
file contents.

```{r files}
files <- httr2::with_mocked_responses(
  function(req) json_response(file_search_json),
  encode_list_files(
    "ENCSR000AAA",
    file_format = "txt",
    quiet = TRUE
  )
)

files
encode_results(files)[, c("file_accession", "file_format", "file_size_pretty", "md5sum")]
```

# Select files

`encode_select_files()` keeps selected rows and records why other rows were not
kept.

```{r select-files}
selected <- encode_select_files(
  files,
  preset = "metadata",
  prefer_default = TRUE,
  explain = FALSE
)

selected
encode_explain_selection(selected)
```

# Preview and dry-run downloads

`encode_preview_download()` checks destination paths and file sizes.
`encode_download(..., dry_run = TRUE)` returns planned download rows without
transferring files.

```{r download}
encode_preview_download(selected, directory = tempdir())

dry_run <- encode_download(
  selected,
  directory = tempdir(),
  dry_run = TRUE
)

encode_results(dry_run)
```

# Record provenance

`encode_manifest()` records selected files, download rows, queries, and citation
metadata. `encode_cite()` returns dataset and file attribution.

```{r manifest}
manifest <- encode_manifest(
  selected,
  include_session = FALSE,
  path = file.path(tempdir(), "encode-manifest.json")
)

names(manifest)
encode_cite(selected, enrich = FALSE)
```

# Read local files

`encode_read()` reads supported files that are already on disk.

```{r read}
txt_path <- tempfile(fileext = ".txt")
writeLines("abc", txt_path)
encode_read(txt_path)

bam_path <- tempfile(fileext = ".bam")
writeBin(charToRaw("placeholder"), bam_path)
encode_read(bam_path)
```
