---
title: "From an ENCODE query to an expression matrix"
author: "Zoheb Khan"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{From an ENCODE query to an expression matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(encodeUtils)
```

# Overview

This vignette retrieves a small set of mouse heart RNA-seq gene-quantification
files from ENCODE and turns compatible tables into count and TPM matrices. The
result is not just downloaded files: it is an auditable hand-off to downstream
analysis, with file metadata, verification results, and ENCODE attribution
recorded in a manifest.

Here, *end-to-end* has a specific meaning. It follows one deliberate set of
records from discovery to local R objects:

| Stage | Question answered | Main result |
|:--|:--|:--|
| Search | Which released experiments could answer the biological question? | `experiments` |
| Review | Which experiment will be used? | `chosen_experiment` |
| List files | What files does that experiment provide? | `files` |
| Select | Which files meet the stated format, output, and assembly criteria? | `selected` |
| Plan and download | Where will files be stored, and did transfer checks pass? | `plan`, then `downloaded` |
| Read | Can the selected tables be safely combined into matrices? | `loaded` |
| Record | What local inputs and software context produced these objects? | `manifest` |

The workflow makes data retrieval and file choice reproducible; it does not
decide whether experiments are biologically comparable or supply a
differential-expression design. Review the experiment metadata, replicates,
life stage, treatment, and processing details before drawing biological
conclusions.

The code below is not evaluated when the vignette is built because it makes a
live ENCODE request and writes files. Run the sections in order in a project
directory. The search result can change as ENCODE records change, so use an
explicit ENCSR accession after you have chosen a dataset for a durable
analysis.

# Goal and starting point

Our immediate goal is deliberately modest: find one *released* mouse heart
RNA-seq experiment, inspect it, select its `mm10` gene-quantification files,
and create matrices when their feature identifiers can be aligned. The
`limit = 1` setting keeps this teaching example small; it does **not** mean
that the first matching experiment is automatically the right experiment for a
scientific analysis.

## 1. Find, inspect, and choose an experiment

`encode_search()` searches metadata only. It does not list or download files.
The default `status = "released"` excludes unreleased records.

```{r search-experiments, eval=FALSE}
experiments <- encode_search(
    organism = "mouse",
    assay = "rna-seq",
    organ = "heart",
    limit = 1
)

experiment_table <- encode_results(experiments)
experiment_table[, c(
    "accession", "assay_title", "biosample_summary", "organism", "status"
)]
```

Check that this row has the tissue, assay, and experimental context you intend
to analyse. With a broader search, filter `experiment_table` using those
metadata and choose an accession explicitly. In this example there should be
one row; save its accession so the next step has an unambiguous input.

```{r choose-experiment, eval=FALSE}
if (nrow(experiment_table) != 1L) {
    stop("Choose one experiment before listing its files.")
}

chosen_experiment <- experiment_table$accession[[1L]]
chosen_experiment
```

For a reproducible analysis, replace the search and choice above with the
recorded accession, for example `chosen_experiment <- "ENCSR..."`.

## 2. List the available file records

An experiment may have raw reads, alignments, signal tracks, and several types
of processed output. `encode_list_files()` retrieves their *metadata*, not
their contents. Start by seeing what is available before applying a filter.

```{r list-files, eval=FALSE}
files <- encode_list_files(chosen_experiment)
file_table <- encode_results(files)

file_table[, c(
    "file_accession", "file_format", "output_type", "assembly",
    "biological_replicates", "preferred_default", "file_size_pretty"
)]
```

At this point, decide what kind of result is appropriate. This tutorial uses
processed gene-quantification tables, not FASTQ reads or alignments. The
`biological_replicates` and `preferred_default` columns are useful checks, but
they do not replace a review of the experiment's design and processing.

## 3. State and audit the file-selection rule

`encode_select_files()` makes the file choice explicit. A preset is a named
format and output-type preference, not a black-box recommendation. Inspect it
before relying on it:

```{r inspect-preset, eval=FALSE}
encode_file_presets("rnaseq_gene_quant")
```

For this preset, the package keeps tabular files and ranks gene-quantification
output types. The additional `assembly = "mm10"` requirement prevents a
mixture of genome assemblies. `preferred_processed` retains every matching
file with the highest-priority output type *within each experiment*; it does
not select a single representative file or discard ties. Add
`prefer_default = TRUE` when you specifically want to restrict to ENCODE files
marked `preferred_default`, if any are available.

```{r select-files, eval=FALSE}
selected <- encode_select_files(
    files,
    preset = "rnaseq_gene_quant",
    assembly = "mm10",
    replicate_policy = "preferred_processed"
)

selected_table <- encode_results(selected)
selected_table[, c(
    "experiment_accession", "file_accession", "file_format",
    "output_type", "assembly", "biological_replicates", "file_size_pretty"
)]

# Why did other file records not qualify, and which settings were applied?
selected$excluded
selected$criteria
```

Do not proceed if `selected_table` is empty or includes files that you would
not analyse. Adjust the biological search, assembly, preset, or selection
policy and inspect this table again.

## 4. Plan the transfer before downloading

First make a dry run. It resolves destination paths and applies the same size
limits as a real transfer, but does not download any bytes. Use a project
directory so that the downloaded files and manifest can travel with the
analysis.

```{r plan-download, eval=FALSE}
plan <- encode_download(
    selected,
    directory = "encode-data",
    dry_run = TRUE
)

plan_table <- encode_results(plan)
plan_table[, c(
    "file_accession", "file_size_pretty", "download_status", "local_path"
)]
```

Review the file count, sizes, and local paths in `plan_table`. By default,
real transfers are bounded to 250 MB per file and 500 MB in total. Files with
unknown sizes are refused. Set `allow_unknown_size = TRUE` only after reviewing
the plan.

## 5. Download and verify the selected files

Remove `dry_run = TRUE` only after the plan looks right. A transfer result has
one row per requested file. A status of `"downloaded"` means this run completed
the transfer; `"exists"` means a matching local file was already present and
passed the requested available checks. `"failed"` must be resolved before
reading the files.

```{r download-files, eval=FALSE}
downloaded <- encode_download(
    selected,
    directory = "encode-data"
)

download_table <- encode_results(downloaded)
download_table[, c(
    "file_accession", "download_status", "local_path",
    "size_verified", "md5_verified", "failure_reason"
)]

ready <- download_table$download_status %in% c("downloaded", "exists")
if (!all(ready)) {
    print(download_table[!ready, c("file_accession", "failure_reason")])
    stop("Resolve failed downloads before reading files.")
}
```

`size_verified` and `md5_verified` are `TRUE` when ENCODE supplied the
corresponding value and that check passed. They are `NA` when no expected value
was available or the check was not requested. Existing files are not
overwritten by default.

## 6. Read compatible quantification tables into R

Pass the completed download table to `encode_read()`. For gene-quantification
tables, it standardizes recognized count columns to `raw_counts` and requests
the expression values you want assembled. Here we request both raw counts and
TPM; raw counts are ordinarily the appropriate input for count-based models,
whereas TPM is useful for descriptive comparisons.

```{r read-files, eval=FALSE}
loaded <- encode_read(
    downloaded,
    values = c("raw_counts", "tpm")
)

# One row per local input file, including the original ENCODE metadata.
loaded$metadata[, c("file_accession", "local_path")]

# Feature annotation and the requested matrices, if safe alignment was possible.
loaded$row_data
loaded$matrices$raw_counts
loaded$matrices$tpm
```

`loaded$data` always retains the individual objects. When every participating
table has a complete, unique feature identifier, `loaded$row_data` and each
matrix share the same rows, and matrix columns are file accessions. If that
condition is not met, `loaded$matrices` is empty rather than aligning rows by
position; inspect the original tables in `loaded$data` before deciding how to
handle them.

## 7. Write a provenance manifest

Finish by writing a machine-readable record alongside the data. Calling
`encode_manifest()` on `loaded` records file metadata and download outcomes,
loaded object classes, ENCODE attribution, package version, and R session
information.

```{r manifest, eval=FALSE}
manifest <- encode_manifest(
    loaded,
    path = "encode-manifest.json"
)

manifest$retrieval
manifest$attribution[, c(
    "experiment_accession", "file_accession", "file_format", "output_type"
)]
```

The in-memory `manifest` and `encode-manifest.json` contain the same
information. Keep the manifest with downstream scripts, matrices, and any
derived results.

# Starting elsewhere

You do not need to begin with a metadata search every time:

- With an ENCSR experiment accession, begin at `encode_list_files()`.
- With ENCFF file accessions, begin at `encode_download()`.
- With downloaded paths or a completed download table, begin at `encode_read()`.

For one local path, `encode_read()` returns the native object for that format.
BED, narrowPeak, and broadPeak files return `GenomicRanges::GRanges` by
default; use `as = "data.frame"` for a plain table. GFF, GTF, BigWig, and
BigBed files use `rtracklayer`, and FASTA files use `Biostrings`, when those
packages are installed. FASTQ, BAM, CRAM, and SAM files remain path objects for
dedicated sequence and alignment readers. Indexed BigWig and BigBed files
require a genomic `region` unless `allow_large = TRUE` is set deliberately.

# Session information

```{r session-info}
sessionInfo()
```
