---
title: "Retrieve ENCODE RNA-seq quantifications"
author: "Zoheb Khan"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Retrieve ENCODE RNA-seq quantifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(encodeUtils)
```

# Introduction

`encodeUtils` turns a defined ENCODE file request into local expression data
with a record of how those files were chosen. It is intended for the retrieval
part of an analysis: finding records, applying explicit file criteria, planning
or performing a bounded download, reading compatible quantification tables, and
recording provenance. It does not decide whether experiments are biologically
comparable or construct a differential-expression design.

This role complements established Bioconductor tools. Use `rtracklayer` to
import a local genomic file, and `AnnotationHub` for curated resources
distributed through Bioconductor. Use `encodeUtils` when the source is the live
ENCODE portal and the search, selection rule, and local files should remain
traceable.

Here, *end-to-end* means one connected sequence:

| Step | Question | Object |
|:--|:--|:--|
| Discover | Which ENCODE experiments are relevant? | `experiments` |
| Select | Which file records meet the stated criteria? | `selected` |
| Plan | What would be downloaded and where? | `plan` |
| Read | Can local tables form aligned matrices? | `loaded` |
| Record | What inputs and software produced the result? | `manifest` |

# Installation

After the package is available through Bioconductor, install it with
BiocManager.

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("encodeUtils")
```

# A reproducible offline walkthrough

Package builds must not depend on the current ENCODE service or transfer remote
files. The rest of this vignette therefore uses two tiny synthetic
gene-quantification tables packaged in `example-data`. The code is evaluated
when the vignette is built. The table metadata uses placeholder URLs only for
the dry-run step; it is not a biological dataset.

```{r example-files}
example_dir <- system.file("example-data", package = "encodeUtils")
files <- utils::read.csv(file.path(example_dir, "files.csv"))
files$local_path <- file.path(example_dir, files$local_name)

files[, c(
    "experiment_accession", "file_accession", "file_format",
    "output_type", "assembly", "biological_replicates"
)]
```

## Select the files

`encode_select_files()` records the selection rule and any excluded files. The
RNA-seq gene-quantification preset requires a tabular gene-quantification
output; `mm10` makes the assembly requirement explicit.

```{r select-files}
selected <- encode_select_files(
    files,
    preset = "rnaseq_gene_quant",
    assembly = "mm10",
    replicate_policy = "preferred_processed",
    quiet = TRUE
)

selected_table <- encode_results(selected)
selected_table[, c(
    "experiment_accession", "file_accession", "file_format",
    "output_type", "assembly"
)]
selected$criteria
```

The example retains both replicate-level quantification files because they
share the highest-priority output type. In a real analysis, inspect
`selected$excluded` and the experiment metadata before continuing.

## Plan a download

A dry run resolves local destinations and applies the usual size limits without
transferring data. It is the right point to check file count and total size.

```{r plan-download}
plan <- encode_download(
    selected,
    directory = file.path(tempdir(), "encode-data"),
    dry_run = TRUE,
    quiet = TRUE
)

encode_results(plan)[, c(
    "file_accession", "file_size_pretty", "download_status", "local_path"
)]
```

## Read local quantification tables

The packaged files already exist locally, so mark them as available and pass
their metadata to `encode_read()`. The loader retains each original table and,
only when feature identifiers are complete and unique, assembles aligned
matrices.

```{r read-files}
local_files <- selected_table
local_files$local_path <- file.path(example_dir, local_files$local_name)
local_files$download_status <- "exists"

loaded <- encode_read(local_files, values = c("raw_counts", "tpm"))
loaded$metadata[, c("file_accession", "local_path")]
loaded$row_data
loaded$matrices$raw_counts
loaded$matrices$tpm
```

For downstream Bioconductor methods, request the same safely aligned data as a
`SummarizedExperiment`. Its assays are the requested expression matrices;
`rowData` holds feature metadata and `colData` holds the ENCODE file metadata.

```{r summarized-experiment}
se <- encode_read(
    local_files,
    values = c("raw_counts", "tpm"),
    as = "SummarizedExperiment"
)

se
SummarizedExperiment::assay(se, "raw_counts")
```

## Record the local analysis

`encode_manifest()` records the local files, matrices, retrieval context, and
session information. This synthetic example suppresses ENCODE attribution
because its file identifiers and URLs are illustrative rather than real.

```{r manifest}
manifest_path <- tempfile(fileext = ".json")
manifest <- encode_manifest(
    loaded,
    include_attribution = FALSE,
    path = manifest_path
)

manifest$retrieval
file.exists(manifest_path)
```

# Live ENCODE retrieval

The following is the same workflow against the live portal. It is deliberately
not evaluated during package builds because both ENCODE records and remote file
transfers can change. Run it in a project directory after reviewing the search
result. Replace the chosen accession with a recorded `ENCSR...` identifier for
a durable analysis.

```{r live-encode, eval=FALSE}
experiments <- encode_search(
    organism = "mouse",
    assay = "rna-seq",
    organ = "heart",
    limit = 1
)
experiment_table <- encode_results(experiments)
chosen_experiment <- experiment_table$accession[[1L]]

files <- encode_list_files(chosen_experiment)
selected <- encode_select_files(
    files,
    preset = "rnaseq_gene_quant",
    assembly = "mm10",
    replicate_policy = "preferred_processed"
)

plan <- encode_download(selected, directory = "encode-data", dry_run = TRUE)
print(encode_results(plan)[, c(
    "file_accession", "file_size_pretty", "download_status", "local_path"
)])

downloaded <- encode_download(selected, directory = "encode-data")
download_table <- encode_results(downloaded)
if (!all(download_table$download_status %in% c("downloaded", "exists"))) {
    stop("Resolve failed downloads before reading files.")
}

loaded <- encode_read(downloaded, values = c("raw_counts", "tpm"))
manifest <- encode_manifest(loaded, path = "encode-manifest.json")
```

Keep `encode-manifest.json` with downstream scripts and derived results. It
records the ENCODE attribution available in the selected file metadata without
making another request.

# Session information

```{r session-info}
sessionInfo()
```
