---
title: "Retrieve ENCODE microRNA-seq quantifications"
author: "Zoheb Khan"
date: "2026-08-06"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Retrieve ENCODE microRNA-seq quantifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 80)
```

```{r library}
library(encodeUtils)
```

# Introduction

`encodeUtils` turns a biological question into an inspectable ENCODE
retrieval. The workflow searches the live ENCODE Portal, applies explicit file
criteria, verifies downloaded files, reads compatible quantification tables,
and records the requests and local artifacts in a manifest. It does not decide
whether experiments are biologically comparable or construct a downstream
statistical design.

The package belongs in Bioconductor because it connects the ENCODE Portal to
standard `GRanges` and `SummarizedExperiment` objects while retaining the
retrieval provenance needed for downstream analyses. It complements format
import through `rtracklayer` and `Biostrings` and general resource catalogs such
as `AnnotationHub`; its narrower role is ENCODE discovery, explicit file
selection, size planning, verified transfer, and manifest creation.

The closest ENCODE-specific Bioconductor resources are `ENCODExplorerData` and
its former companion software package, `ENCODExplorer`, which was removed in
Bioconductor 3.15. `ENCODExplorerData` provides ENCODE metadata snapshots
through `AnnotationHub` and helpers for regenerating those tables. In contrast,
`encodeUtils` queries the current Portal for each retrieval and carries explicit
file choices through size planning, verification, standard Bioconductor
containers, and a provenance manifest.

This vignette executes the complete workflow against the public ENCODE service.
The discovery search is deliberately dynamic because portal contents and search
ordering change over time. Downstream steps use one named released experiment
and two named files so that the transferred inputs are explicit. The files are
mouse-heart microRNA count tables of approximately 61 KB each.

| Step | Question | Object |
|:--|:--|:--|
| Discover | Which ENCODE experiments are currently relevant? | `discovery` |
| List | Which files belong to the pinned experiment? | `files` |
| Select | Which exact released files meet the stated criteria? | `selected` |
| Download | Were planned and observed sizes verified? | `downloaded` |
| Read | What tables, feature metadata, and matrix were loaded? | `loaded` |
| Convert | Are rows and columns aligned in a Bioconductor container? | `se` |
| Record | Which requests and files produced the result? | `manifest` |

# Installation

After the package is available through Bioconductor, install it with
BiocManager.

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("encodeUtils")
```

# Discover current mouse-heart microRNA-seq experiments

The first query is a real biological discovery search. Its result is a current
view of the portal rather than a frozen claim about which experiment will sort
first in the future.

```{r discover}
discovery <- encode_search(
    organism = "mouse",
    assay = "microRNA-seq",
    organ = "heart",
    limit = 5,
    metadata = "full",
    quiet = TRUE
)

discovery_table <- as.data.frame(encode_results(discovery))
discovery_table[, c("accession", "assay_title", "status"), drop = FALSE]
```

The returned experiments all match the search, but they differ in strain, age,
sex, and other biological characteristics. Inspect those characteristics before
choosing experiments for an analysis. For this retrieval demonstration, the
downstream inputs are pinned rather than inferred from the first search row.

Experiment `ENCSR523CTA` contains two biological replicates from heart tissue
of male, 8-to-10-month-old 5xFAD/CAST mice. The tutorial uses its two released
`mm10` microRNA quantification files.

```{r pinned-inputs}
experiment_accession <- "ENCSR523CTA"
file_accessions <- c("ENCFF859GWB", "ENCFF838WBE")
```

# List and select the pinned files

Listing retrieves current metadata for all released `mm10` microRNA
quantification TSVs attached to the experiment. It does not transfer file
contents.

```{r list-files}
files <- encode_list_files(
    experiment_accession,
    file_format = "tsv",
    output_type = "microRNA quantifications",
    assembly = "mm10",
    status = "released",
    limit = "all",
    metadata = "full",
    quiet = TRUE
)

file_table <- as.data.frame(encode_results(files))
file_table[, c(
    "file_accession", "biological_replicates", "file_size",
    "preferred_default"
), drop = FALSE]
```

The selection names both file accessions explicitly. Format, output type,
assembly, released status, and replicate-level requirements remain independent
checks rather than being implied by the accessions.

```{r select-files}
selected <- encode_select_files(
    files,
    file_accession = file_accessions,
    file_format = "tsv",
    output_type = "microRNA quantifications",
    assembly = "mm10",
    status = "released",
    replicate_policy = "replicate_level",
    quiet = TRUE
)

selected_table <- as.data.frame(encode_results(selected))
selected_table[, c(
    "file_accession", "biological_replicates", "file_size"
), drop = FALSE]
```

The effective criteria are retained with the selected files and later copied
into the manifest.

```{r selection-criteria}
selected$criteria[c(
    "file_accession", "file_format", "output_type", "assembly", "status",
    "replicate_policy"
)]
```

# Plan and perform a verified download

The dry run confirms the filenames and ENCODE-reported sizes before any bytes
are transferred. Both the per-file and total budgets are intentionally close to
the expected transfer, so a substantial upstream file change fails instead of
silently increasing build resources. These are planning limits based on Portal
metadata; the observed size and MD5 are checked after transfer.

```{r plan-download}
download_directory <- file.path(tempdir(), "encodeUtils-live-vignette")

plan <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB",
    dry_run = TRUE,
    quiet = TRUE
)

plan_table <- as.data.frame(encode_results(plan))
data.frame(
    file_accession = plan_table$file_accession,
    size = plan_table$file_size_pretty,
    status = plan_table$download_status,
    local_file = basename(plan_table$local_path)
)
```

The evaluated transfer writes to the session temporary directory. Each file is
installed at its destination only after its expected size and MD5 checksum have
been verified.

```{r download}
downloaded <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB",
    quiet = TRUE
)

download_table <- as.data.frame(encode_results(downloaded))
download_table[, c(
    "file_accession", "downloaded_size", "size_verified", "md5_verified"
), drop = FALSE]
```

```{r verify-download, include=FALSE}
stopifnot(
    all(download_table$download_status %in% c("downloaded", "exists")),
    all(download_table$size_verified %in% TRUE),
    all(download_table$md5_verified %in% TRUE)
)
```

# Read and inspect the quantifications

The two TSV files contain source gene identifiers and raw counts. They do not
contain TPM values, so this tutorial requests only `raw_counts`; it does not
construct or imply a TPM matrix.

`encode_read()` retains each simplified source table and constructs a matrix
only because the tables have the same complete feature set and compatible
ENCODE metadata. The files use `gene_id`, so that column is requested explicitly
for row names.

```{r read-files}
loaded <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id"
)

names(loaded)
```

The `metadata` component retains the full ENCODE and download metadata. This
compact view uses basenames so temporary build paths do not dominate the page.

```{r inspect-loaded-metadata}
loaded_file_summary <- data.frame(
    file_accession = loaded$metadata$file_accession,
    replicate = loaded$metadata$biological_replicates,
    download_status = loaded$metadata$download_status,
    local_file = basename(loaded$metadata$local_path)
)
loaded_file_summary
```

The `data` component retains one simplified source table per downloaded file.

```{r inspect-loaded-data}
data.frame(
    file_accession = names(loaded$data),
    rows = vapply(loaded$data, nrow, integer(1L)),
    columns = vapply(loaded$data, ncol, integer(1L)),
    row.names = NULL
)
```

The `row_data` component describes the features in the combined matrix. Only
the first six rows are displayed.

```{r inspect-row-data}
data.frame(
    gene_id = unname(loaded$row_data$gene_id[seq_len(6L)])
)
```

The `matrices` component contains exactly the requested value type. Its columns
are the two ENCODE file accessions.

```{r inspect-matrices}
names(loaded$matrices)
```

```{r inspect-raw-counts}
loaded$matrices$raw_counts[seq_len(6L), , drop = FALSE]
```

# Create a SummarizedExperiment

The same verified files can be returned in a standard Bioconductor container.
File metadata are aligned to assay columns and feature metadata are aligned to
assay rows.

```{r summarized-experiment}
se <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id",
    as = "SummarizedExperiment"
)

se
```

The displayed column metadata confirm which file produced each assay column.

```{r inspect-column-data}
column_data <- as.data.frame(SummarizedExperiment::colData(se))
row.names(column_data) <- NULL
column_data[, c(
    "file_accession", "biological_replicates", "output_type", "assembly"
), drop = FALSE]
```

The package-level metadata retain the source query, every metadata request, and
the effective file-selection criteria after conversion to the standard
container.

```{r inspect-se-provenance}
se_provenance <- S4Vectors::metadata(se)$encodeUtils
data.frame(
    requests = length(se_provenance$requests),
    filter_rows = nrow(se_provenance$filters),
    selection_fields = length(se_provenance$selection_criteria)
)
```

# Record the retrieval

The manifest captures the ENCODE metadata-request history for the pinned file
workflow, effective selection criteria, file accessions and URLs, expected and
observed checksums, download times, matrix dimensions, package version, and R
session information.

```{r create-manifest}
manifest_path <- file.path(tempdir(), "encodeUtils-live-manifest.json")
manifest <- encode_manifest(loaded, path = manifest_path)
```

The retrieval summary records when the manifest and its source metadata were
created. The complete query URL remains in `manifest$retrieval$query_url` and in
the JSON file.

```{r inspect-retrieval}
manifest$retrieval[c("created_at", "retrieved_at")]
```

Each ENCODE metadata request is retained with its URL and query. This compact
view confirms the request role, HTTP status, and retrieval time.

```{r inspect-requests}
manifest$requests[[1L]][c("role", "status_code", "retrieved_at")]
```

The manifest also records the matrix dimensions and both the expected and
observed checksums.

```{r inspect-manifest-matrix}
matrix_summary <- manifest$matrices
row.names(matrix_summary) <- NULL
matrix_summary
```

```{r inspect-manifest-files}
manifest$files[, c(
    "file_accession", "md5sum", "observed_md5", "downloaded_at"
), drop = FALSE]
```

Finally, confirm that the JSON manifest was written.

```{r inspect-manifest-path}
file.exists(manifest_path)
```

Stable accessions and checksums make the transferred inputs auditable, but a
live service can still be unavailable or retire a record. `encodeUtils` uses
bounded retries and timeouts and reports the failing URL. Keeping the manifest
with downstream scripts records exactly what was available when the workflow
ran.

# Further reading

- ENCODE Project Consortium. An integrated encyclopedia of DNA elements in the
    human genome. *Nature*. 2012;489:57-74.
- Kagda MS et al. Data navigation on the ENCODE portal. *Nature
    Communications*. 2025;16:9592.
- [ENCODE REST API](https://www.encodeproject.org/help/rest-api/)
- [ENCODE citation guidance](https://www.encodeproject.org/help/citing-encode/)
- [ENCODExplorerData](https://bioconductor.org/packages/ENCODExplorerData/)
- [Bioconductor removed
    packages](https://bioconductor.org/about/removed-packages/)
- [encodeUtils package website](https://zohebkhan1.github.io/encodeUtils/)

# Session information

```{r session-info}
sessionInfo()
```
