---
title: "Retrieve ENCODE microRNA-seq quantifications"
author: "Zoheb Khan"
date: "2026-08-07"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Retrieve ENCODE microRNA-seq quantifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

options(width = 80)
```

# Introduction

The [ENCODE Portal](https://www.encodeproject.org/) is a public repository of
experiment records, biosample descriptions, data files, and metadata from the
ENCODE Project. ENCODE provides programmatic access to this information through
a REST API. The API returns data in JSON format.

`encodeUtils` provides an R interface for searching ENCODE metadata, selecting
and verifying files, reading supported formats, and recording retrieval
provenance. It converts JSON responses from the API into R and Bioconductor
objects.

This vignette shows how to retrieve replicate-level raw microRNA counts from
male 5xFAD mouse heart tissue. It begins with biological search criteria, and
each step derives its input from the output of the preceding step.

# ENCODE API disclaimer

The following rate-limit disclaimer is reproduced directly from the
[ENCODE REST API documentation](https://www.encodeproject.org/help/rest-api/):

> Use of the programmatic API is limited to 10 GET requests/sec from any single
> user, group, company or lab.
>
> Abuse of this will result in denial of access by IP block

# Installation

```{r installation, eval=FALSE}
BiocManager::install("encodeUtils")
```

Install the development version with:

```{r development-installation, eval=FALSE}
pak::pak("ZohebKhan1/encodeUtils")
```

# Find the experiment

Begin with the biological criteria. The `search` argument looks for `5xFAD` in
ENCODE metadata. The remaining arguments restrict the results to male mouse
heart microRNA-seq experiments. To investigate a different question, change
`organism`, `assay`, `organ`, `sex`, and `search`. `limit` sets the maximum
number of results and does not filter by biology.

```{r find-experiment}
library(encodeUtils)

discovery <- encode_search(
    organism = "mouse",
    assay = "microRNA-seq",
    organ = "heart",
    sex = "male",
    search = "5xFAD",
    limit = 5
)

experiment_table <- encode_results(discovery)

experiment_table[, c(
    "accession", "assay_title", "biosample_summary", "life_stage_age", "sex",
    "file_count", "status"
), drop = FALSE]
```

The search identifies a released microRNA-seq experiment from male 5xFAD/CAST
mouse heart tissue. The current query returns one experiment. If a search
returns several experiments, compare their assay, biosample, age, sex, and
status fields, then refine the search arguments before continuing. The check
below stops the workflow unless the live search returns exactly one experiment,
preventing the code from selecting the first result without explanation.

```{r choose-experiment}
stopifnot(nrow(experiment_table) == 1L)

experiment_accession <- experiment_table$accession[[1L]]
experiment_accession
```

# List the experiment's files

Use the accession obtained above to retrieve metadata for every released file
associated with the experiment. This step does not download file contents.

```{r list-files}
files <- encode_list_files(
    experiment_accession
)

file_table <- encode_results(files)

file_inventory <- data.frame(
    accession = file_table$file_accession,
    replicate = file_table$biological_replicates,
    format = file_table$file_format,
    output = file_table$output_type,
    assembly = file_table$assembly,
    size = file_table$file_size_pretty
)

knitr::kable(file_inventory)
```

# Select replicate-level files

`file_format` identifies the storage format, while `output_type` identifies the
data product. To build a count matrix, select TSV files containing `microRNA
quantifications`. These processed files use the `mm10` mouse assembly, and
`replicate_level` retains a separate file for each biological replicate. To
retrieve another data product, inspect the inventory and change the relevant
selection arguments.

```{r select-files}
selected <- encode_select_files(
    files,
    file_format = "tsv",
    output_type = "microRNA quantifications",
    assembly = "mm10",
    replicate_policy = "replicate_level"
)

selected_table <- encode_results(selected)

selected_display <- data.frame(
    accession = selected_table$file_accession,
    replicate = selected_table$biological_replicates,
    format = selected_table$file_format,
    output = selected_table$output_type,
    assembly = selected_table$assembly,
    size = selected_table$file_size_pretty
)

knitr::kable(selected_display)
```

The selection narrows the inventory from `r nrow(file_table)` files to
`r nrow(selected_table)` quantification files. The remaining steps use these
accessions, and the selection object retains the criteria used to choose them.

```{r inspect-selection}
selected$criteria[c(
    "file_format", "output_type", "assembly", "status", "replicate_policy"
)]
```

# Preview the download

Create a download plan before transferring either file. The plan determines the
destination paths and compares the file sizes reported by ENCODE with the
configured limits, but it does not download the files. The limits guard against
an unexpectedly large transfer if ENCODE metadata changes.

Each selected file is about 60 KB, so the 100 KB per-file and 200 KB total
limits provide modest headroom above the expected 120 KB transfer.

The vignette uses a temporary directory so package builds do not leave data in
the source tree. Files under `tempdir()` are temporary. For an analysis, use a
persistent, project-relative directory instead.

```{r plan-download}
download_directory <- file.path(tempdir(), "encodeUtils-live-vignette")

plan <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB",
    dry_run = TRUE
)

plan
```

Confirm the file accessions, destination paths, and expected total size before
downloading.

# Download and verify the files

Run the same request without `dry_run`. `encode_download()` first writes each
transfer to a temporary `.part` file, then verifies its size and MD5 checksum
against ENCODE metadata before moving it to the requested destination.

```{r download}
downloaded <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB"
)

downloaded
```

The following checks require every local file to have a verified size and
checksum.

```{r verify-download}
download_table <- as.data.frame(encode_results(downloaded))

stopifnot(
    all(download_table$download_status %in% c("downloaded", "exists")),
    all(download_table$size_verified %in% TRUE),
    all(download_table$md5_verified %in% TRUE)
)

download_table[, c(
    "file_accession", "download_status", "size_verified", "md5_verified"
), drop = FALSE]
```

# Read the count tables

The selected microRNA quantification files are four-column tables in HTSeq
output format. `encodeUtils` stores the raw-count column in a matrix named
`raw_counts`. Both tables contain complete, unique `gene_id` values, which
`encode_read()` uses to align their rows. The matrix columns correspond to the
file accessions shown earlier.

```{r read-files}
count_data <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id"
)

count_data
```

`count_data` contains the source tables, file and feature metadata, and the
aligned raw-count matrix.

Inspect the first six rows of the count matrix.

```{r inspect-counts}
count_data$matrices$raw_counts[seq_len(6L), , drop = FALSE]
```

# Write a manifest

Create a JSON manifest for the completed retrieval. The manifest records the
file-listing request stored in `count_data`, filters, selection criteria,
downloaded files, checksums, matrix dimensions, and R session information. The
earlier experiment search remains in `discovery` and is not part of this
manifest.

```{r create-manifest}
manifest_path <- file.path(tempdir(), "encodeUtils-live-manifest.json")

manifest <- encode_manifest(count_data, path = manifest_path)

data.frame(
    manifest = basename(attr(manifest, "path")),
    files = nrow(manifest$files),
    requests = length(manifest$requests),
    matrices = nrow(manifest$matrices)
)
```

This temporary manifest exists only for the package build. For an analysis,
write it to a persistent, project-relative path.

Inspect the file records stored in the manifest.

```{r inspect-manifest-files}
manifest$files[, c(
    "file_accession", "md5sum", "observed_md5", "downloaded_at"
), drop = FALSE]
```

# Optional: create a SummarizedExperiment

To work with a Bioconductor container instead, read the same verified local
files into a `SummarizedExperiment`. The assay stores raw counts, `rowData()`
stores feature identifiers, and `colData()` stores the ENCODE file metadata.

```{r summarized-experiment}
se <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id",
    as = "SummarizedExperiment"
)

se
```

The column metadata links each matrix column to its ENCODE file and biological
replicate.

```{r inspect-column-data}
column_data <- as.data.frame(SummarizedExperiment::colData(se))

column_data[, c(
    "file_accession", "biological_replicates", "output_type", "assembly"
), drop = FALSE]
```

# References

- ENCODE Project Consortium. An integrated encyclopedia of DNA elements in the
  human genome. *Nature*. 2012;489:57-74.
- Kagda MS et al. Data navigation on the ENCODE portal. *Nature
  Communications*. 2025;16:9592.
- [ENCODE REST API](https://www.encodeproject.org/help/rest-api/)
- [ENCODE citation guidance](https://www.encodeproject.org/help/citing-encode/)

# Session information

```{r session-info}
sessionInfo()
```
