---
title: "Get started"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(encodeUtils)
```

# Introduction

`encodeUtils` provides a small R-native workflow for finding ENCODE datasets,
inspecting the files attached to those datasets, planning safe downloads,
reading supported local files, and recording provenance. The package is intended
for scripted high-throughput sequencing workflows where users need to move from
an ENCODE accession or search result to files that can be used in R.

The package complements the ENCODE web portal and REST API. It does not replace
Bioconductor import tools; instead, it keeps ENCODE discovery and download
bookkeeping in R and delegates supported genomic file imports to packages such
as `rtracklayer` and `Biostrings` when those packages are installed. BED-like
interval files are returned as `GRanges` by default when `GenomicRanges` and
`IRanges` are available, and raw sequencing or alignment files are returned as
paths for use with downstream tools such as `ShortRead`, `Rsamtools`, or
`GenomicAlignments`.

# Installation

Install the current package from GitHub:

```{r install-github, eval=FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}

pak::pak("ZohebKhan1/encodeUtils")
```

After a Bioconductor release becomes available, install it with:

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("encodeUtils")
```

# Workflow

Most analyses follow the same sequence:

1. Search for experiments, or start from known ENCODE accessions.
2. List files attached to selected experiments.
3. Select the file accessions or file types you want.
4. Check the download with `dry_run = TRUE`.
5. Download files when the plan is acceptable.
6. Read supported local files into R.
7. Save a manifest for the files you used.

The evaluated examples below use small local ENCODE-like records so the vignette
is reproducible on the Bioconductor build system. A live ENCODE example is shown
later as non-evaluated code because public web resources can be temporarily
unavailable or change over time.

# Select files

Create a small file table like the one returned by `encode_list_files()`.

```{r example-files}
example_files <- data.frame(
  experiment_accession = "ENCSR000AAA",
  file_accession = c("ENCFF000AAA", "ENCFF000AAB", "ENCFF000AAC"),
  file_format = c("tsv", "tsv", "bigWig"),
  output_type = c(
    "gene quantifications",
    "gene quantifications",
    "signal p-value"
  ),
  assembly = c("mm10", "mm10", "mm10"),
  status = c("released", "released", "released"),
  file_size = c(120, 140, 5000000),
  md5sum = NA_character_,
  href = c(
    "/files/ENCFF000AAA/@@download/ENCFF000AAA.tsv",
    "/files/ENCFF000AAB/@@download/ENCFF000AAB.tsv",
    "/files/ENCFF000AAC/@@download/ENCFF000AAC.bigWig"
  ),
  stringsAsFactors = FALSE
)

selected <- encode_select_files(
  example_files,
  preset = "rnaseq_gene_quant",
  assembly = "mm10"
)

encode_results(selected)
```

# Check a download

`encode_download(dry_run = TRUE)` builds the destination paths and checks known
sizes without transferring any file contents.

```{r dry-run}
planned <- encode_download(
  selected,
  directory = tempdir(),
  dry_run = TRUE,
  quiet = TRUE
)

encode_results(planned)[, c("file_accession", "download_status", "local_path")]
```

# Read a local file

After files are downloaded, `encode_read()` can read supported local formats. The
example below writes a tiny gene-quantification table to a temporary file and
then reads it through the same metadata shape returned by `encode_download()`.

```{r read-local}
quant_path <- tempfile(fileext = ".tsv")
writeLines(
  c(
    "gene_id\tTPM\tFPKM\traw_counts",
    "ENSMUSG00000000001\t4.5\t2.1\t11",
    "ENSMUSG00000000028\t8.2\t4.4\t27"
  ),
  quant_path
)

downloaded <- encode_results(planned)
downloaded$local_path <- quant_path
downloaded$download_status <- "downloaded"
downloaded$organism <- "Mus musculus"

loaded <- encode_read(
  downloaded,
  values = c("raw_counts", "TPM"),
  row_names = "none"
)

loaded$metadata
loaded$raw_counts
loaded$tpm
```

For loaded collections, the canonical components are `metadata`, `data`,
`matrices`, and `by_experiment`. The `files`, `raw_counts`, and `tpm` elements
are convenience aliases for `metadata`, `matrices$raw_counts`, and
`matrices$TPM` when those objects are available. If code should handle one-file
and many-file reads with the same object shape, use
`encode_read(downloaded[1, ], as_collection = TRUE)`.

Human and mouse RNA quantification tables are annotated with gene symbols,
Ensembl IDs, and Entrez IDs when `org.Hs.eg.db` or `org.Mm.eg.db` is installed.
If the relevant annotation package is unavailable, `encodeUtils` keeps the
original identifiers and continues without that optional annotation step.

# Save provenance

`encode_manifest()` records the package version, retrieval metadata, file table,
and ENCODE attribution fields when they are available.

```{r manifest}
manifest <- encode_manifest(downloaded, include_session = FALSE)
names(manifest)
```

# Live ENCODE use

The same workflow can be run against the live ENCODE Portal. Start from a narrow
query or a known experiment accession, inspect the returned rows, and perform a
dry run before transferring files.

```{r live-encode-example, eval=FALSE}
rna_experiments <- encode_search(
  type = "Experiment",
  organism = "mouse",
  assay = "rna-seq",
  organ = "heart",
  status = "released",
  limit = 5
)

rna_files <- encode_list_files(
  rna_experiments,
  file_format = "tsv",
  output_type = "gene quantifications",
  assembly = "mm10",
  allow_many = TRUE
)

rna_table <- encode_results(rna_files)
rna_table <- rna_table[seq_len(min(2, nrow(rna_table))), ]

encode_download(
  rna_table,
  directory = tempdir(),
  dry_run = TRUE
)
```

# Return objects

`encodeUtils` returns compact S3 objects that print as small tables. Use
`encode_results()` when ordinary data-frame operations are needed.

The default print methods intentionally show a compact subset of columns. For
file tables, the user-facing columns are the file accession, experiment
accession, assay, organism, biosample, file format, output type, assembly, file
size, and status. The full returned table keeps the broader ENCODE metadata for
filtering, joins, and provenance; retrieve it with `encode_results()`.

Common columns for downstream code:

- `file_accession` identifies the ENCODE file;
- `experiment_accession` identifies the parent experiment;
- `organism` and `biosample_term_name` identify species and sample;
- `file_format`, `output_type`, and `assembly` are the main file-selection
  fields;
- `file_size`, `md5sum`, `download_url`, and `local_path` support download
  planning, verification, and provenance.

For local file reading:

- a single local path or one downloaded-file row returns the native object for
  that file, whereas a multi-row downloaded-file table returns an
  `encode_loaded_files` collection;
- TSV and CSV files are returned as data frames;
- ENCODE gene-quantification tables are normalized to common identifier and
  expression columns by default, and `simplify_quant = FALSE` preserves raw
  downloaded columns;
- BED, narrowPeak, and broadPeak files are returned as `GRanges` when the
  optional genomic reader stack can parse the file; ENCODE peak files with extra
  nonstandard columns may fall back to a plain table unless `as = "GRanges"` is
  requested;
- FASTA files use `Biostrings` when it is installed;
- indexed genomic formats such as bigWig and bigBed use `rtracklayer` when it is
  installed and the requested read is small or region-limited;
- FASTQ and alignment files return an `encode_local_file` path object by default
  because read processing is normally performed by dedicated Bioconductor or
  command-line workflows;
- large or unsupported files return an `encode_local_file` path object unless
  `unsupported = "error"` is requested.

# Session information

```{r session-info}
sessionInfo()
```
