---
title: "Get started"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r library}
library(encodeUtils)
```

# Workflow

Most analyses follow the same sequence:

1. Search for experiments.
2. List files attached to those experiments.
3. Select the file accessions you want.
4. Check the download with `dry_run = TRUE`.
5. Download the files.
6. Read supported files into R.
7. Save a manifest for the files you used.

`encode_results()` extracts the main table from any result object when you want
to filter, save, or inspect it with ordinary R code.

# Bulk RNA-seq

Search for mouse heart total RNA-seq experiments.

```{r rnaseq-search}
rna_experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart total RNA-seq",
  status = "released",
  limit = 10
)

encode_results(rna_experiments)
```

List gene-quantification TSV files for those experiments.

```{r rnaseq-files}
rna_files <- encode_list_files(
  rna_experiments,
  file_format = "tsv",
  output_type = "gene quantifications",
  assembly = "mm10"
)

encode_results(rna_files)
```

Choose the exact files to use. File accessions are the safest way to avoid
accidentally downloading the first rows of a broad search.

```{r rnaseq-select}
rna_file_ids <- c("ENCFF090VKE", "ENCFF260OJQ")
```

Check the transfer before downloading.

```{r rnaseq-dry-run}
rna_dry_run <- encode_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = "data/encode/rna-seq",
  dry_run = TRUE
)

encode_results(rna_dry_run)
```

Download and read the files.

```{r rnaseq-download}
rna_downloaded <- encode_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = "data/encode/rna-seq"
)

rna_loaded <- encode_read(rna_downloaded)
```

The loaded object keeps the data and metadata together.

```{r rnaseq-loaded}
rna_loaded$metadata
rna_loaded$data$ENCFF090VKE
rna_loaded$matrices$expected_count
rna_loaded$matrices$TPM
```

# ATAC-seq

Search for mouse heart ATAC-seq experiments.

```{r atac-search}
atac_experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart ATAC-seq",
  status = "released",
  limit = 10
)

encode_results(atac_experiments)
```

List peak and signal files.

```{r atac-files}
atac_files <- encode_list_files(
  atac_experiments,
  file_format = c("bed", "bigWig"),
  assembly = "mm10"
)

encode_results(atac_files)
```

Select peak files, then check the download.

```{r atac-select}
atac_selected <- encode_select_files(
  atac_files,
  preset = "atacseq_peaks",
  assembly = "mm10"
)

atac_dry_run <- encode_download(
  atac_selected,
  directory = "data/encode/atac-seq",
  dry_run = TRUE
)
```

# ChIP-seq

Search for mouse heart ChIP-seq experiments.

```{r chip-search}
chip_experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart ChIP-seq",
  status = "released",
  limit = 10
)

encode_results(chip_experiments)
```

For ChIP-seq, inspect the `target` and `control_type` columns. Histone and TF
experiments should show the target. Control experiments usually have
`control_type = "input library"` and no target.

List peak files and select IDR peaks.

```{r chip-files}
chip_files <- encode_list_files(
  chip_experiments,
  file_format = "bed",
  assembly = "mm10"
)

chip_selected <- encode_select_files(
  chip_files,
  preset = "chipseq_idr_peaks",
  assembly = "mm10"
)

encode_results(chip_selected)
```

Check the download.

```{r chip-dry-run}
chip_dry_run <- encode_download(
  chip_selected,
  directory = "data/encode/chip-seq",
  dry_run = TRUE
)

encode_results(chip_dry_run)
```

# Provenance

Use a manifest to save the query, selected files, downloaded paths, and ENCODE
attribution metadata for the files used in an analysis.

```{r manifest}
manifest <- encode_manifest(
  rna_downloaded,
  include_session = FALSE,
  path = "data/encode/rna-seq/encode-manifest.json"
)

names(manifest)
```
