---
title: "Retrieve ENCODE microRNA-seq quantifications"
author: "Zoheb Khan"
date: "2026-08-07"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Retrieve ENCODE microRNA-seq quantifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

options(width = 80)
```

# Introduction

The ENCODE Portal is a public database of experiments, biosamples, files, and
associated metadata produced by the ENCODE Project. The Portal provides an
HTTP-based REST API that returns its records as JSON.

`encodeUtils` uses this API to search ENCODE, select files from returned
metadata, download and verify those files, read supported formats, and record
retrieval provenance.

This vignette retrieves replicate-level raw microRNA counts for male 5xFAD
mouse heart tissue. It starts with biological search terms, then uses each
result as the input to the next step.

# Warning

Use of the programmatic API is limited to 10 GET requests/sec from any single
user, group, company or lab.

Abuse of this will result in denial of access by IP block

# Installation

```{r installation, eval=FALSE}
BiocManager::install("encodeUtils")
```

Install the development version with:

```{r development-installation, eval=FALSE}
pak::pak("ZohebKhan1/encodeUtils")
```

# Find the experiment

Start with the biological criteria. The `search` argument matches `5xFAD` in
the ENCODE record. The remaining arguments restrict the results to male mouse
heart microRNA-seq experiments. For another question, change `organism`,
`assay`, `organ`, `sex`, and `search`. `limit` controls the maximum number of
returned rows; it is not a biological filter.

```{r find-experiment}
library(encodeUtils)

discovery <- encode_search(
    organism = "mouse",
    assay = "microRNA-seq",
    organ = "heart",
    sex = "male",
    search = "5xFAD",
    limit = 5
)

experiment_table <- encode_results(discovery)

experiment_table[, c(
    "accession", "assay_title", "biosample_summary", "life_stage_age", "sex",
    "file_count", "status"
), drop = FALSE]
```

The result is a released male mouse-heart microRNA-seq experiment whose
biosample summary contains `5xFAD/CAST`. The current query returns one
experiment. If a search returns several rows, compare their assay, biosample,
age, sex, and status fields, then refine the search arguments before continuing.
The check below prevents the workflow from silently taking the first row if the
Portal results change.

```{r choose-experiment}
stopifnot(nrow(experiment_table) == 1L)

experiment_accession <- experiment_table$accession[[1L]]
experiment_accession
```

# List the experiment's files

Use the returned experiment accession to list all of its released files. This
call retrieves file metadata; it does not download file contents.

```{r list-files}
files <- encode_list_files(
    experiment_accession
)

file_table <- encode_results(files)

file_inventory <- data.frame(
    accession = file_table$file_accession,
    replicate = file_table$biological_replicates,
    format = file_table$file_format,
    output = file_table$output_type,
    assembly = file_table$assembly,
    size = file_table$file_size_pretty
)

knitr::kable(file_inventory)
```

# Select replicate-level files

`file_format` describes how a file is encoded; `output_type` describes the data
product it contains. The goal is a count matrix, so select the TSV rows whose
output type is `microRNA quantifications`. `mm10` is the mouse assembly used for
these processed files, and `replicate_level` keeps separate biological
replicates. For another data product, inspect the inventory and change the
relevant selection arguments.

```{r select-files}
selected <- encode_select_files(
    files,
    file_format = "tsv",
    output_type = "microRNA quantifications",
    assembly = "mm10",
    replicate_policy = "replicate_level"
)

selected_table <- encode_results(selected)

selected_display <- data.frame(
    accession = selected_table$file_accession,
    replicate = selected_table$biological_replicates,
    format = selected_table$file_format,
    output = selected_table$output_type,
    assembly = selected_table$assembly,
    size = selected_table$file_size_pretty
)

knitr::kable(selected_display)
```

The selection reduces the `r nrow(file_table)`-file inventory to
`r nrow(selected_table)` quantification files. These are the file accessions
used in the remaining steps. The selection object also records the applied
criteria.

```{r inspect-selection}
selected$criteria[c(
    "file_format", "output_type", "assembly", "status", "replicate_policy"
)]
```

# Preview the download

Create a download plan before transferring either file. The plan resolves the
local paths and compares ENCODE-reported sizes with the configured limits; it
does not inspect remote file contents. The limits prevent an unexpected large
transfer if the Portal records change.

Each selected file is about 60 KB, so the 100 KB per-file and 200 KB total
limits leave modest headroom above the expected 120 KB transfer.

The vignette uses a temporary directory so package builds do not leave files in
the project. Files under `tempdir()` are disposable. Use a persistent
project-relative directory when retaining data for an analysis.

```{r plan-download}
download_directory <- file.path(tempdir(), "encodeUtils-live-vignette")

plan <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB",
    dry_run = TRUE
)

plan
```

Review the planned accessions, paths, and total size before downloading.

# Download and verify the files

Run the same request without `dry_run`. `encode_download()` writes each
transfer to a partial file, then checks its observed size and MD5 checksum
before keeping it.

```{r download}
downloaded <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB"
)

downloaded
```

The following check confirms that both rows represent verified local files.

```{r verify-download}
download_table <- as.data.frame(encode_results(downloaded))

stopifnot(
    all(download_table$download_status %in% c("downloaded", "exists")),
    all(download_table$size_verified %in% TRUE),
    all(download_table$md5_verified %in% TRUE)
)

download_table[, c(
    "file_accession", "download_status", "size_verified", "md5_verified"
), drop = FALSE]
```

# Read the count tables

The selected microRNA quantification files are four-column HTSeq-style tables.
`encodeUtils` maps the second count column to `raw_counts`. The tables share a
complete, unique `gene_id` field, so `encode_read()` uses it to align rows and
creates one matrix whose columns are the file accessions shown earlier.

```{r read-files}
count_data <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id"
)

count_data
```

`count_data` contains the individual file tables, their metadata, feature
metadata, and the aligned raw-count matrix.

Inspect the first six rows of the count matrix.

```{r inspect-counts}
count_data$matrices$raw_counts[seq_len(6L), , drop = FALSE]
```

# Write a manifest

Write a JSON manifest for the completed file retrieval. The manifest records
the file-listing request carried by `count_data`, filters, selection criteria,
downloaded files, checksums, matrix dimensions, and R session. The earlier
experiment search remains in `discovery`; it is not included in this manifest.

```{r create-manifest}
manifest_path <- file.path(tempdir(), "encodeUtils-live-manifest.json")

manifest <- encode_manifest(count_data, path = manifest_path)

data.frame(
    manifest = basename(attr(manifest, "path")),
    files = nrow(manifest$files),
    requests = length(manifest$requests),
    matrices = nrow(manifest$matrices)
)
```

The temporary manifest is used only for this package build. Use a persistent
project-relative path when retaining the manifest with an analysis.

Inspect the file records stored in the manifest.

```{r inspect-manifest-files}
manifest$files[, c(
    "file_accession", "md5sum", "observed_md5", "downloaded_at"
), drop = FALSE]
```

# Optional: create a SummarizedExperiment

Instead of working with the list and matrix result, request the same verified
local files as a `SummarizedExperiment`. The assay stores raw counts,
`rowData()` stores feature identifiers, and `colData()` stores the ENCODE file
metadata.

```{r summarized-experiment}
se <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id",
    as = "SummarizedExperiment"
)

se
```

The column metadata connects each matrix column to the file and biological
replicate obtained from the file-listing step.

```{r inspect-column-data}
column_data <- as.data.frame(SummarizedExperiment::colData(se))

column_data[, c(
    "file_accession", "biological_replicates", "output_type", "assembly"
), drop = FALSE]
```

# References

- ENCODE Project Consortium. An integrated encyclopedia of DNA elements in the
  human genome. *Nature*. 2012;489:57-74.
- Kagda MS et al. Data navigation on the ENCODE portal. *Nature
  Communications*. 2025;16:9592.
- [ENCODE REST API](https://www.encodeproject.org/help/rest-api/)
- [ENCODE citation guidance](https://www.encodeproject.org/help/citing-encode/)

# Session information

```{r session-info}
sessionInfo()
```
