---
title: "Query and retrieve ENCODE files"
author: "Zoheb Khan"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Query and retrieve ENCODE files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(encodeUtils)
```

# Introduction

`encodeUtils` connects ENCODE metadata discovery to explicit file selection,
bounded transfers, local readers, and reproducibility manifests. Search and
file-list results retain the query URL and complete flattened metadata. The
examples below are local and deterministic so the vignette can be built without
network access.

The package fills a current Bioconductor access gap left by the retired
`ENCODExplorer`: it queries current ENCODE Portal records on demand, preserves
request and attribution metadata, and applies explicit transfer limits and
verification. It complements rather than replaces format-specific
Bioconductor infrastructure. BED, GFF, GTF, BigWig, and BigBed files use
`rtracklayer`; FASTA files use `Biostrings`; BED-like tables return
`GenomicRanges::GRanges` by default; and sequence-read and alignment files stay
as paths for dedicated readers such as `ShortRead`, `Rsamtools`, or
`GenomicAlignments`.

# Installation

Install `encodeUtils` from Bioconductor with:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("encodeUtils")
```

# Select files

Start with a small table shaped like the output of `encode_list_files()`.

```{r example-files}
files <- data.frame(
  experiment_accession = "ENCSR000AAA",
  file_accession = c("ENCFF000AAA", "ENCFF000AAB", "ENCFF000AAC"),
  file_format = c("tsv", "tsv", "bigWig"),
  output_type = c(
    "gene quantifications",
    "gene quantifications",
    "signal p-value"
  ),
  assembly = "mm10",
  status = "released",
  file_size = c(100, 120, 5000000),
  md5sum = NA_character_,
  href = c(
    "/files/ENCFF000AAA/@@download/ENCFF000AAA.tsv",
    "/files/ENCFF000AAB/@@download/ENCFF000AAB.tsv",
    "/files/ENCFF000AAC/@@download/ENCFF000AAC.bigWig"
  ),
  stringsAsFactors = FALSE
)

encode_file_presets()

selected <- encode_select_files(
  files,
  preset = "rnaseq_gene_quant",
  assembly = "mm10"
)

encode_results(selected)
```

The selected-file object also retains excluded rows and the criteria used.

```{r selection-audit}
names(selected)
selected$criteria$preset
```

# Plan a download

Use `dry_run = TRUE` to resolve destination paths and check known size limits
without making a network request.

```{r download-plan}
planned <- encode_download(
  selected,
  directory = tempdir(),
  dry_run = TRUE,
  quiet = TRUE
)

encode_results(planned)[, c(
  "file_accession", "file_size_pretty", "download_status", "local_path"
)]
```

For a real transfer, run the same call without `dry_run = TRUE`.
`directory = NULL` uses the persistent package cache; `directory = tempdir()`
is appropriate for transient work.

# Read downloaded tables

The next chunk creates two small quantification files and attaches their local
paths to the planned metadata. A downloaded-file table always returns the same
collection shape, even when it has one row.

```{r read-local}
quant_paths <- file.path(tempdir(), c("ENCFF000AAA.tsv", "ENCFF000AAB.tsv"))
writeLines(
  c(
    "gene_id\traw_counts\tTPM",
    "Gata4\t11\t4.5",
    "Tbx5\t27\t8.2"
  ),
  quant_paths[[1L]]
)
writeLines(
  c(
    "gene_id\traw_counts\tTPM",
    "Gata4\t14\t5.1",
    "Tbx5\t31\t9.0"
  ),
  quant_paths[[2L]]
)

downloaded <- encode_results(planned)
downloaded$local_path <- quant_paths
downloaded$download_status <- "downloaded"

loaded <- encode_read(
  downloaded,
  values = c("raw_counts", "tpm")
)

names(loaded)
loaded$metadata[, c("file_accession", "local_path")]
loaded$row_data
loaded$matrices$raw_counts
loaded$matrices$tpm
```

The `metadata` component is the complete input file table. `data` contains one
native object per file. `row_data` and each numeric matrix share row names, and
matrix columns are file accessions. Use `simplify_quant = FALSE` when the exact
downloaded columns must be preserved.

# Record provenance

`encode_manifest()` captures the package version, retrieval context, files,
loaded object classes, and ENCODE attribution fields. Set `path` to write the
same manifest as JSON.

```{r manifest}
manifest <- encode_manifest(loaded, include_session = FALSE)
names(manifest)
manifest$attribution[, c(
  "experiment_accession", "file_accession", "file_format", "output_type"
)]
```

# Live ENCODE workflow

The same sequence can start from a live query. This block runs only in an
interactive R session, keeping package builds deterministic.

```{r live-workflow}
if (interactive()) {
  experiments <- encode_search(
    type = "Experiment",
    organism = "mouse",
    assay = "rna-seq",
    organ = "heart",
    limit = 5
  )

  files <- encode_list_files(
    experiments,
    file_format = "tsv",
    output_type = "gene quantifications",
    assembly = "mm10"
  )

  selected <- encode_select_files(files, preset = "rnaseq_gene_quant")
  encode_download(selected, directory = tempdir(), dry_run = TRUE)
}
```

# Session information

```{r session-info}
sessionInfo()
```
