---
title: "Get started"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`encodeUtils` is built around a deliberate ENCODE workflow:

1. Search for experiments or files.
2. Inspect the returned metadata.
3. List files attached to the experiments you care about.
4. Select the exact files you want.
5. Preview the download size and destination paths.
6. Run a dry-run before a real download.
7. Save a manifest and citation table for reproducibility.

The package is metadata-first. Searching, listing, selecting, previewing, and
creating manifests do not download data files. A real transfer happens only when
you call `encode_download()` with `dry_run = FALSE`.

# Mental model

ENCODE has many object types. The two most important for everyday use are:

- **Experiment**: a dataset page on the ENCODE website. One experiment usually
  has many files.
- **File**: one downloadable data file inside an experiment, such as a FASTQ,
  BED, bigWig, BAM, or quantification table.

In most workflows you search for experiments first, then list the files attached
to those experiments.

```{r load-package}
library(encodeUtils)
```

# Bulk RNA-seq

This example searches for mouse heart total RNA-seq experiments, then narrows to
gene-quantification TSV files.

```{r rnaseq-search}
rna_experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart total RNA-seq",
  status = "released",
  limit = 10,
  metadata = "full"
)

rna_experiments
```

Use `encode_results()` when you want the ordinary data frame.

```{r rnaseq-results}
encode_results(rna_experiments)
```

Now list file metadata for those experiments. This still does not download any
files.

```{r rnaseq-files}
rna_files <- encode_list_files(
  rna_experiments,
  file_format = "tsv",
  output_type = "gene quantifications",
  assembly = "mm10",
  metadata = "full"
)

rna_files
```

For a small test, choose file accessions from the printed table. Replace these
IDs with files from your own result.

```{r rnaseq-preview}
rna_file_ids <- c("ENCFF260OJQ", "ENCFF090VKE")

rna_plan <- encode_preview_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = tempdir()
)

rna_plan
```

Run a dry-run before a real transfer. A dry-run returns planned download rows but
does not download the files.

```{r rnaseq-dry-run}
rna_dry_run <- encode_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = tempdir(),
  dry_run = TRUE
)

encode_results(rna_dry_run)
```

When the plan is correct, remove `dry_run = TRUE` or set `dry_run = FALSE`.

```{r rnaseq-download}
rna_downloaded <- encode_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = "data/encode/rna-seq",
  dry_run = FALSE
)
```

# ATAC-seq

ATAC-seq workflows usually start at experiments and then select peak files.

```{r atac-search}
atac_experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart ATAC-seq",
  status = "released",
  limit = 10,
  metadata = "full"
)

atac_experiments
```

List candidate peak files.

```{r atac-files}
atac_files <- encode_list_files(
  atac_experiments,
  file_format = "bed",
  assembly = "mm10",
  metadata = "full"
)

atac_files
```

Apply the ATAC-seq peak preset. The selection object keeps both selected files
and excluded files, so the decision can be audited.

```{r atac-select}
atac_selected <- encode_select_files(
  atac_files,
  preset = "atacseq_peaks",
  assembly = "mm10"
)

atac_selected
encode_explain_selection(atac_selected)
```

Preview the download before transfer.

```{r atac-preview}
atac_plan <- encode_preview_download(
  atac_selected,
  directory = tempdir()
)

atac_plan
```

# ChIP-seq

For ChIP-seq, the target and control information matter. The compact search
table includes target information when ENCODE provides it.

```{r chip-search}
chip_experiments <- encode_search(
  type = "Experiment",
  search = "mouse heart H3K27ac ChIP-seq",
  status = "released",
  limit = 10,
  metadata = "full"
)

chip_experiments
```

List peak files for the experiments.

```{r chip-files}
chip_peak_files <- encode_list_files(
  chip_experiments,
  file_format = "bed",
  assembly = "mm10",
  metadata = "full"
)

chip_peak_files
```

Select ChIP-seq peak files with the preset.

```{r chip-select-peaks}
chip_peaks <- encode_select_files(
  chip_peak_files,
  preset = "chipseq_idr_peaks",
  assembly = "mm10"
)

chip_peaks
encode_explain_selection(chip_peaks)
```

For signal tracks, list bigWig files and use the signal preset.

```{r chip-signal}
chip_signal_files <- encode_list_files(
  chip_experiments,
  file_format = "bigWig",
  assembly = "mm10",
  metadata = "full"
)

chip_signal <- encode_select_files(
  chip_signal_files,
  preset = "chipseq_signal_bigwig",
  assembly = "mm10"
)

chip_signal
```

Preview before downloading.

```{r chip-preview}
chip_plan <- encode_preview_download(
  chip_peaks,
  directory = tempdir()
)

chip_plan
```

# Manifest and citation

After selecting files, save enough information to reproduce the choice later.

```{r manifest}
manifest <- encode_manifest(
  rna_dry_run,
  include_session = FALSE
)

encode_write_manifest(
  manifest,
  path = file.path(tempdir(), "encode-rna-manifest.json")
)
```

Create a citation/provenance table for the datasets and files used.

```{r cite}
encode_cite(
  rna_dry_run,
  enrich = "auto"
)
```

# Practical notes

- Use `limit = 10` or another small number while learning.
- Use `encode_count()` only as a preflight for very broad queries.
- Use `encode_preview_download()` before real downloads.
- Use `file_accession = c("ENCFF...", "ENCFF...")` when you want specific
  files from a table.
- Use `tempdir()` for testing. Use a deliberate project directory for real work.

# Session information

```{r session-info, eval=TRUE}
utils::sessionInfo()
```
