---
title: "Retrieve ENCODE microRNA-seq quantifications"
author: "Zoheb Khan"
date: "2026-08-07"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Retrieve ENCODE microRNA-seq quantifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 80)
```

```{r library}
library(encodeUtils)
```

# Workflow

This vignette retrieves two released microRNA count tables from ENCODE
experiment `ENCSR523CTA`, verifies both files, reads their raw counts, creates a
`SummarizedExperiment`, and writes a retrieval manifest. The initial discovery
query is live; the download uses pinned accessions.

# Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("encodeUtils")
```

# Find experiments

Search current mouse-heart microRNA-seq experiments.

```{r discover}
discovery <- encode_search(
    organism = "mouse",
    assay = "microRNA-seq",
    organ = "heart",
    limit = 5,
    metadata = "full",
    quiet = TRUE
)

encode_results(discovery)
```

Use experiment `ENCSR523CTA` and its two released `mm10` replicate-level
quantification files.

```{r pinned-inputs}
experiment_accession <- "ENCSR523CTA"
file_accessions <- c("ENCFF859GWB", "ENCFF838WBE")
```

# Select files

Retrieve file metadata without downloading file contents.

```{r list-files}
files <- encode_list_files(
    experiment_accession,
    file_format = "tsv",
    output_type = "microRNA quantifications",
    assembly = "mm10",
    status = "released",
    limit = "all",
    metadata = "full",
    quiet = TRUE
)

encode_results(files)[, c(
    "file_accession", "biological_replicates", "file_size_pretty",
    "preferred_default"
), drop = FALSE]
```

Select the two named files and require replicate-level output.

```{r select-files}
selected <- encode_select_files(
    files,
    file_accession = file_accessions,
    file_format = "tsv",
    output_type = "microRNA quantifications",
    assembly = "mm10",
    status = "released",
    replicate_policy = "replicate_level",
    quiet = TRUE
)

selected
selected$excluded
```

```{r selection-criteria}
selected$criteria[c(
    "file_accession", "file_format", "output_type", "assembly", "status",
    "replicate_policy"
)]
```

# Preview and download

Check the ENCODE-reported sizes without transferring either file.

```{r plan-download}
download_directory <- file.path(tempdir(), "encodeUtils-live-vignette")

plan <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB",
    dry_run = TRUE,
    quiet = TRUE
)

plan
```

Download both files and verify their observed sizes and MD5 checksums.

```{r download}
downloaded <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB",
    quiet = TRUE
)

downloaded
```

```{r verify-download, include=FALSE}
download_table <- as.data.frame(encode_results(downloaded))
stopifnot(
    all(download_table$download_status %in% c("downloaded", "exists")),
    all(download_table$size_verified %in% TRUE),
    all(download_table$md5_verified %in% TRUE)
)
```

# Read raw counts

The files contain `gene_id` and raw counts, but not TPM values.

```{r read-files}
loaded <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id"
)

loaded
```

```{r inspect-loaded-files}
data.frame(
    file_accession = names(loaded$data),
    replicate = loaded$metadata$biological_replicates,
    rows = vapply(loaded$data, nrow, integer(1L)),
    columns = vapply(loaded$data, ncol, integer(1L)),
    row.names = NULL
)
```

```{r inspect-raw-counts}
loaded$matrices$raw_counts[seq_len(6L), , drop = FALSE]
```

# Create a SummarizedExperiment

```{r summarized-experiment}
se <- encode_read(
    downloaded,
    values = "raw_counts",
    row_names = "gene_id",
    as = "SummarizedExperiment"
)

se
```

```{r inspect-column-data}
column_data <- as.data.frame(SummarizedExperiment::colData(se))
row.names(column_data) <- NULL
column_data[, c(
    "file_accession", "biological_replicates", "output_type", "assembly"
), drop = FALSE]
```

```{r inspect-se-provenance}
se_provenance <- S4Vectors::metadata(se)$encodeUtils
data.frame(
    requests = length(se_provenance$requests),
    filter_rows = nrow(se_provenance$filters),
    selection_fields = length(se_provenance$selection_criteria)
)
```

# Write a manifest

```{r create-manifest}
manifest_path <- file.path(tempdir(), "encodeUtils-live-manifest.json")
manifest <- encode_manifest(loaded, path = manifest_path)

data.frame(
    manifest = basename(attr(manifest, "path")),
    files = nrow(manifest$files),
    requests = length(manifest$requests),
    matrices = nrow(manifest$matrices)
)
```

```{r inspect-manifest-files}
manifest$files[, c(
    "file_accession", "md5sum", "observed_md5", "downloaded_at"
), drop = FALSE]
```

The live steps use finite timeouts and retries. A completed download is kept
only after its size and MD5 checksum pass verification.

# References

- ENCODE Project Consortium. An integrated encyclopedia of DNA elements in the
    human genome. *Nature*. 2012;489:57-74.
- Kagda MS et al. Data navigation on the ENCODE portal. *Nature
    Communications*. 2025;16:9592.
- [ENCODE REST API](https://www.encodeproject.org/help/rest-api/)
- [ENCODE citation guidance](https://www.encodeproject.org/help/citing-encode/)

# Session information

```{r session-info}
sessionInfo()
```
