---
title: "Get started"
author: "Zoheb Khan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r helpers, include=FALSE}
library(encodeUtils)

json_response <- function(body) {
  httr2::response(
    200,
    headers = "Content-Type: application/json",
    body = charToRaw(body)
  )
}

rows <- function(x) {
  as.data.frame(encode_results(x), stringsAsFactors = FALSE)
}

show_columns <- function(x, columns) {
  x <- as.data.frame(x, stringsAsFactors = FALSE)
  x[, intersect(columns, names(x)), drop = FALSE]
}

show_experiments <- function(x) {
  show_columns(
    rows(x),
    c(
      "accession", "assay_title", "target", "control_type", "organism",
      "biosample_term_name", "biosample_type", "sex", "biosample_summary",
      "lab", "project", "file_count", "date_released", "status"
    )
  )
}

show_files <- function(x) {
  show_columns(
    rows(x),
    c(
      "experiment_accession", "file_accession", "assay_title", "target",
      "biosample_term_name", "biosample_type", "sex", "file_format",
      "file_type", "output_type", "assembly", "analysis_accession",
      "file_size_pretty", "status"
    )
  )
}

show_plan <- function(x) {
  show_columns(
    encode_results(x),
    c(
      "experiment_accession", "file_accession", "file_format",
      "output_type", "assembly", "file_size_pretty", "local_path"
    )
  )
}

show_explain <- function(x) {
  show_columns(
    encode_explain_selection(x),
    c("file_accession", "experiment_accession", "decision", "reason")
  )
}

download_marker <- paste0(intToUtf8(64), intToUtf8(64), "download")

rna_experiment_json <- paste0(
  '{"@graph":[',
  '{"accession":"ENCSR083OKX","@id":"/experiments/ENCSR083OKX/",',
  '"@type":["Experiment"],"assay_title":"total RNA-seq",',
  '"organism":{"scientific_name":"Mus musculus"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"female postnatal (4 days) strain B6CASTF1/J heart tissue",',
  '"lab":{"title":"Barbara Wold, Caltech"},"award":{"project":"ENCODE"},',
  '"date_released":"2021-08-18","files":[1,2],"status":"released"},',
  '{"accession":"ENCSR967IKT","@id":"/experiments/ENCSR967IKT/",',
  '"@type":["Experiment"],"assay_title":"total RNA-seq",',
  '"organism":{"scientific_name":"Mus musculus"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"female adult (8-10 months) strain 5xFAD/CAST heart tissue",',
  '"lab":{"title":"Barbara Wold, Caltech"},"award":{"project":"ENCODE"},',
  '"date_released":"2022-02-23","files":[1,2],"status":"released"}',
  '],"total":80}'
)

rna_file_json <- paste0(
  '{"@graph":[',
  '{"accession":"ENCFF090VKE","@id":"/files/ENCFF090VKE/",',
  '"dataset":"/experiments/ENCSR083OKX/","assay_title":"total RNA-seq",',
  '"simple_biosample_summary":"female postnatal (4 days) strain B6CASTF1/J heart tissue",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"tsv","file_type":"tsv","output_type":"gene quantifications",',
  '"assembly":"mm10","file_size":11890884,',
  '"md5sum":"11111111111111111111111111111111",',
  '"href":"/files/ENCFF090VKE/', download_marker, '/ENCFF090VKE.tsv",',
  '"analyses":["/analyses/ENCAN155LXI/"],"status":"released"},',
  '{"accession":"ENCFF260OJQ","@id":"/files/ENCFF260OJQ/",',
  '"dataset":"/experiments/ENCSR967IKT/","assay_title":"total RNA-seq",',
  '"simple_biosample_summary":"female adult (8-10 months) strain 5xFAD/CAST heart tissue",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"tsv","file_type":"tsv","output_type":"gene quantifications",',
  '"assembly":"mm10","file_size":11900950,',
  '"md5sum":"22222222222222222222222222222222",',
  '"href":"/files/ENCFF260OJQ/', download_marker, '/ENCFF260OJQ.tsv",',
  '"analyses":["/analyses/ENCAN631YVW/"],"status":"released"}',
  '],"total":2}'
)

atac_experiment_json <- paste0(
  '{"@graph":[',
  '{"accession":"ENCSR000ATC","@id":"/experiments/ENCSR000ATC/",',
  '"@type":["Experiment"],"assay_title":"ATAC-seq",',
  '"organism":{"scientific_name":"Mus musculus"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"mouse heart tissue embryo (14.5 days)",',
  '"lab":{"title":"ENCODE DCC"},"award":{"project":"ENCODE"},',
  '"date_released":"2021-02-01","files":[1,2],"status":"released"}',
  '],"total":12}'
)

atac_file_json <- paste0(
  '{"@graph":[',
  '{"accession":"ENCFF000ATC","@id":"/files/ENCFF000ATC/",',
  '"dataset":"/experiments/ENCSR000ATC/","assay_title":"ATAC-seq",',
  '"simple_biosample_summary":"mouse heart tissue embryo (14.5 days)",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"bed","file_type":"bed narrowPeak",',
  '"output_type":"IDR thresholded peaks","assembly":"mm10",',
  '"file_size":882134,"md5sum":"33333333333333333333333333333333",',
  '"href":"/files/ENCFF000ATC/', download_marker, '/ENCFF000ATC.bed.gz",',
  '"preferred_default":true,"status":"released"},',
  '{"accession":"ENCFF000ATD","@id":"/files/ENCFF000ATD/",',
  '"dataset":"/experiments/ENCSR000ATC/","assay_title":"ATAC-seq",',
  '"simple_biosample_summary":"mouse heart tissue embryo (14.5 days)",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"bigWig","output_type":"signal p-value","assembly":"mm10",',
  '"file_size":4822000,"md5sum":"44444444444444444444444444444444",',
  '"href":"/files/ENCFF000ATD/', download_marker, '/ENCFF000ATD.bigWig",',
  '"status":"released"}',
  '],"total":2}'
)

chip_experiment_json <- paste0(
  '{"@graph":[',
  '{"accession":"ENCSR582SPN","@id":"/experiments/ENCSR582SPN/",',
  '"@type":["Experiment"],"assay_title":"Histone ChIP-seq",',
  '"target":{"label":"H3K27ac-mouse"},',
  '"control_type":null,',
  '"organism":{"scientific_name":"Mus musculus"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"mouse heart tissue embryo (10.5 days)",',
  '"lab":{"title":"Bing Ren, UCSD"},"award":{"project":"ENCODE"},',
  '"date_released":"2016-07-19","files":[1,2],"status":"released"},',
  '{"accession":"ENCSR077QRY","@id":"/experiments/ENCSR077QRY/",',
  '"@type":["Experiment"],"assay_title":"Control ChIP-seq",',
  '"control_type":"input library",',
  '"organism":{"scientific_name":"Mus musculus"},',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"biosample_summary":"mouse heart tissue embryo (10.5 days)",',
  '"lab":{"title":"Bing Ren, UCSD"},"award":{"project":"ENCODE"},',
  '"date_released":"2015-06-23","files":[1],"status":"released"}',
  '],"total":5}'
)

chip_peak_json <- paste0(
  '{"@graph":[',
  '{"accession":"ENCFF000CHP","@id":"/files/ENCFF000CHP/",',
  '"dataset":"/experiments/ENCSR582SPN/","assay_title":"Histone ChIP-seq",',
  '"target":{"label":"H3K27ac-mouse"},',
  '"simple_biosample_summary":"mouse heart tissue embryo (10.5 days)",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"bed","file_type":"bed narrowPeak",',
  '"output_type":"optimal IDR thresholded peaks","assembly":"mm10",',
  '"file_size":2190442,"md5sum":"55555555555555555555555555555555",',
  '"href":"/files/ENCFF000CHP/', download_marker, '/ENCFF000CHP.bed.gz",',
  '"preferred_default":true,"status":"released"},',
  '{"accession":"ENCFF000CHS","@id":"/files/ENCFF000CHS/",',
  '"dataset":"/experiments/ENCSR582SPN/","assay_title":"Histone ChIP-seq",',
  '"target":{"label":"H3K27ac-mouse"},',
  '"simple_biosample_summary":"mouse heart tissue embryo (10.5 days)",',
  '"biosample_ontology":{"term_name":"heart","classification":"tissue"},',
  '"file_format":"bigWig","output_type":"fold change over control",',
  '"assembly":"mm10","file_size":6720010,',
  '"md5sum":"66666666666666666666666666666666",',
  '"href":"/files/ENCFF000CHS/', download_marker, '/ENCFF000CHS.bigWig",',
  '"status":"released"}',
  '],"total":2}'
)
```

# Overview

ENCODE records are searched before any files are downloaded. A typical analysis
starts with experiments, lists the files attached to those experiments, selects
the files to use, checks the planned download, and records provenance.

Two ENCODE record types are used most often:

- **Experiment**: a dataset page on the ENCODE website.
- **File**: one downloadable file attached to an experiment.

# Bulk RNA-seq

Search for mouse heart total RNA-seq experiments.

```{r rnaseq-search}
rna_experiments <- httr2::with_mocked_responses(
  function(req) json_response(rna_experiment_json),
  encode_search(
    type = "Experiment",
    search = "mouse heart total RNA-seq",
    status = "released",
    limit = 10,
    quiet = TRUE
  )
)
```

The search returns matching ENCODE experiments.

```{r rnaseq-results, echo=FALSE}
knitr::kable(
  show_experiments(rna_experiments),
  row.names = FALSE
)
```

List gene-quantification TSV files for the returned experiments.

```{r rnaseq-files}
rna_files <- httr2::with_mocked_responses(
  function(req) json_response(rna_file_json),
  encode_list_files(
    rna_experiments,
    file_format = "tsv",
    output_type = "gene quantifications",
    assembly = "mm10",
    quiet = TRUE
  )
)
```

```{r rnaseq-file-table, echo=FALSE}
knitr::kable(
  show_files(rna_files),
  row.names = FALSE
)
```

Preview two specific files before transfer.

```{r rnaseq-preview}
rna_file_ids <- c("ENCFF090VKE", "ENCFF260OJQ")

rna_plan <- encode_preview_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = tempdir(),
  quiet = TRUE
)
```

```{r rnaseq-preview-table, echo=FALSE}
knitr::kable(
  show_plan(rna_plan),
  row.names = FALSE
)
```

Run a dry-run to return the planned download rows without downloading files.

```{r rnaseq-dry-run}
rna_dry_run <- encode_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = tempdir(),
  dry_run = TRUE,
  quiet = TRUE
)
```

```{r rnaseq-dry-run-table, echo=FALSE}
knitr::kable(
  show_plan(rna_dry_run),
  row.names = FALSE
)
```

# ATAC-seq

Search for mouse heart ATAC-seq experiments.

```{r atac-search}
atac_experiments <- httr2::with_mocked_responses(
  function(req) json_response(atac_experiment_json),
  encode_search(
    type = "Experiment",
    search = "mouse heart ATAC-seq",
    status = "released",
    limit = 10,
    quiet = TRUE
  )
)
```

```{r atac-experiment-table, echo=FALSE}
knitr::kable(
  show_experiments(atac_experiments),
  row.names = FALSE
)
```

List candidate peak and signal files.

```{r atac-files}
atac_files <- httr2::with_mocked_responses(
  function(req) json_response(atac_file_json),
  encode_list_files(
    atac_experiments,
    file_format = c("bed", "bigWig"),
    assembly = "mm10",
    quiet = TRUE
  )
)
```

```{r atac-file-table, echo=FALSE}
knitr::kable(
  show_files(atac_files),
  row.names = FALSE
)
```

Select ATAC-seq peak files.

```{r atac-select}
atac_selected <- encode_select_files(
  atac_files,
  preset = "atacseq_peaks",
  assembly = "mm10",
  explain = FALSE
)
```

```{r atac-select-table, echo=FALSE}
knitr::kable(
  show_files(atac_selected),
  row.names = FALSE
)
knitr::kable(
  show_explain(atac_selected),
  row.names = FALSE
)
```

```{r atac-preview}
atac_plan <- encode_preview_download(
  atac_selected,
  directory = tempdir(),
  quiet = TRUE
)
```

```{r atac-preview-table, echo=FALSE}
knitr::kable(
  show_plan(atac_plan),
  row.names = FALSE
)
```

# ChIP-seq

For ChIP-seq, target and control information are part of the experiment table.

```{r chip-search}
chip_experiments <- httr2::with_mocked_responses(
  function(req) json_response(chip_experiment_json),
  encode_search(
    type = "Experiment",
    search = "mouse heart H3K27ac ChIP-seq",
    status = "released",
    limit = 10,
    quiet = TRUE
  )
)
```

```{r chip-experiment-table, echo=FALSE}
knitr::kable(
  show_experiments(chip_experiments),
  row.names = FALSE
)
```

List peak and signal files.

```{r chip-files}
chip_files <- httr2::with_mocked_responses(
  function(req) json_response(chip_peak_json),
  encode_list_files(
    chip_experiments,
    file_format = c("bed", "bigWig"),
    assembly = "mm10",
    quiet = TRUE
  )
)
```

```{r chip-file-table, echo=FALSE}
knitr::kable(
  show_files(chip_files),
  row.names = FALSE
)
```

Select peak files.

```{r chip-select-peaks}
chip_peaks <- encode_select_files(
  chip_files,
  preset = "chipseq_idr_peaks",
  assembly = "mm10",
  explain = FALSE
)
```

```{r chip-select-peaks-table, echo=FALSE}
knitr::kable(
  show_files(chip_peaks),
  row.names = FALSE
)
knitr::kable(
  show_explain(chip_peaks),
  row.names = FALSE
)
```

Select signal tracks from the same file table.

```{r chip-signal}
chip_signal <- encode_select_files(
  chip_files,
  preset = "chipseq_signal_bigwig",
  assembly = "mm10",
  explain = FALSE
)
```

```{r chip-signal-table, echo=FALSE}
knitr::kable(
  show_files(chip_signal),
  row.names = FALSE
)
```

# Manifest and citation

Save a manifest after files have been selected or downloaded.

```{r manifest}
manifest <- encode_manifest(
  rna_dry_run,
  include_session = FALSE,
  path = file.path(tempdir(), "encode-rna-manifest.json")
)
```

```{r manifest-table, echo=FALSE}
knitr::kable(
  data.frame(section = names(manifest)),
  row.names = FALSE
)
```

Create a citation table for the datasets and files used.

```{r cite, echo=FALSE}
knitr::kable(
  show_columns(
    encode_cite(rna_dry_run, enrich = FALSE),
    c("accession", "type", "url", "dataset", "file_accession")
  ),
  row.names = FALSE
)
```

# Real downloads

The examples above use `tempdir()` and `dry_run = TRUE`. For a real transfer,
use a project directory and set `dry_run = FALSE`.

```r
downloaded <- encode_download(
  rna_files,
  file_accession = rna_file_ids,
  directory = "data/encode/rna-seq",
  dry_run = FALSE
)
```
