---
title: "Retrieve ENCODE microRNA-seq quantifications"
author: "Zoheb Khan"
date: "2026-08-07"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Retrieve ENCODE microRNA-seq quantifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

The [ENCODE Portal](https://www.encodeproject.org/) is a public repository of
experiment records, biosample descriptions, data files, and metadata from the
ENCODE Project. ENCODE provides programmatic access to this information through
a REST API.

`encodeUtils` searches ENCODE metadata, selects and verifies files, reads local
formats, and records retrieval provenance. This vignette retrieves
replicate-level raw microRNA counts from male 5xFAD mouse heart tissue.

ENCODE data are common inputs to Bioconductor functional genomics workflows,
but the Portal exposes nested metadata and heterogeneous file types that can
require separate querying, selection, and provenance steps. The archived
[ENCODExplorer](https://bioconductor.org/packages/3.12/bioc/html/ENCODExplorer.html)
package combined compiled ENCODE metadata with API, download, and SQLite
helpers. `encodeUtils` instead focuses on live Portal searches, explicit file
selection and verification, recorded request and file provenance, and supported
outputs as `GenomicRanges::GRanges` or `SummarizedExperiment` objects.

# ENCODE API disclaimer

ENCODE limits programmatic access to 10 GET requests per second from one user,
group, company, or lab, and warns that abuse can result in an IP block.
`encodeUtils` throttles requests to five per second.

See the [ENCODE data-use policy](https://www.encodeproject.org/help/citing-encode/)
for use and citation guidance.

# Installation

```{r installation, eval=FALSE}
install.packages("BiocManager")
BiocManager::install("encodeUtils")
```

Install the development version with:

```{r development-installation, eval=FALSE}
install.packages("pak")
pak::pak("ZohebKhan1/encodeUtils")
```

# Find the experiment

`encode_search()` accepts exact ENCODE Portal fields in `filters`. This avoids
assuming that a convenient assay label represents every related assay. The
free-text term locates 5xFAD records, while the named fields restrict organism,
assay, tissue, and sex.

```{r find-experiment}
library(encodeUtils)

discovery <- encode_search(
    filters = list(
        "replicates.library.biosample.organism.scientific_name" = "Mus musculus",
        assay_title = "microRNA-seq",
        "biosample_ontology.organ_slims" = "heart",
        "replicates.library.biosample.sex" = "male"
    ),
    search = "5xFAD",
    limit = 5
)

experiment_table <- encode_results(discovery)

experiment_table[c(
    "accession", "assay_title", "biosample_summary", "life_stage_age", "sex",
    "file_count", "status"
)]
```

The query currently identifies one released experiment. If several experiments
match, compare their assay, biosample, age, sex, and status fields before
continuing.

# List the quantification files

File format, output type, assembly, and status are server-side filters in
`encode_list_files()`. The returned table retains the exact request and does not
download file contents.

```{r list-files}
files <- encode_list_files(
    experiment_table$accession[[1L]],
    file_format = "tsv",
    output_type = "microRNA quantifications",
    assembly = "mm10"
)

file_table <- encode_results(files)
file_table[c(
    "file_accession", "biological_replicates", "file_size_pretty", "status"
)]
```

# Select replicate-level files

`replicate_level` retains separate biological-replicate files. File-field
filters are not repeated at this stage; use the file query above or subset the
table with ordinary R operations.

```{r select-files}
selected <- encode_select_files(
    files,
    replicate_policy = "replicate_level"
)

selected_table <- encode_results(selected)
selected_table[c(
    "file_accession", "biological_replicates", "file_size_pretty"
)]
```

# Preview and download

Downloads are uncapped by default. This example supplies opt-in limits because
the two expected files are small. `dry_run = TRUE` determines destination paths
and checks ENCODE-reported sizes without transferring data.

```{r plan-download}
download_directory <- file.path(tempdir(), "encodeUtils-live-vignette")

plan <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB",
    dry_run = TRUE
)

encode_results(plan)[c(
    "file_accession", "download_status", "file_size_pretty", "local_path"
)]
```

Run the same request without `dry_run`. Transfers use temporary `.part` files.
Size is checked when ENCODE reports it, and MD5 is checked when ENCODE supplies
a checksum.

```{r download}
downloaded <- encode_download(
    selected,
    directory = download_directory,
    max_file_size = "100KB",
    max_total_size = "200KB"
)

download_table <- encode_results(downloaded)

stopifnot(
    all(download_table$download_status %in% c("downloaded", "exists")),
    all(download_table$size_verified %in% TRUE),
    all(download_table$md5_verified %in% TRUE)
)
```

# Read and prepare the count tables

The downloaded quantification files are headerless four-column HTSeq tables.
`encode_read_all()` reads them verbatim, so the file-specific `header = FALSE`
choice is explicit. `encode_prepare_quant()` then names the identifier and raw
count columns and removes HTSeq summary rows. Existing columns are retained.

```{r read-files}
count_data <- encode_read_all(downloaded, header = FALSE)
count_data <- encode_prepare_quant(count_data)

head(count_data$data[[1L]])
```

# Create a SummarizedExperiment

Matrix assembly is also explicit. Every table must contain a complete, unique
`gene_id`, the same feature set, and a numeric `raw_counts` column. Incompatible
inputs produce an error instead of an empty matrix with a hidden reason.

```{r summarized-experiment}
se <- encode_as_summarized_experiment(
    count_data,
    feature_id = "gene_id",
    assays = "raw_counts"
)

print(se)
```

The assay columns correspond to ENCODE file accessions, and `colData()` retains
the complete file metadata.

```{r inspect-counts}
head(SummarizedExperiment::assay(se, "raw_counts"), 5L)

as.data.frame(SummarizedExperiment::colData(se))[c(
    "file_accession", "biological_replicates", "output_type", "assembly"
)]
```

# Write a manifest

The manifest records the file-listing request retained by `count_data`, active
filters, selection criteria, downloaded files, checksums, loaded-object classes,
and R session information. It uses local metadata and makes no API request.

```{r create-manifest}
manifest_path <- file.path(tempdir(), "encodeUtils-live-manifest.json")
manifest <- encode_manifest(count_data, path = manifest_path)

manifest$files[c(
    "file_accession", "md5sum", "observed_md5", "downloaded_at"
)]
```

# Other inputs and formats

BED, narrowPeak, and broadPeak files return `GenomicRanges::GRanges` by default.
GFF, GTF, BigWig, and BigBed use `rtracklayer` when installed, while FASTA uses
`Biostrings`. Large, unsupported, FASTQ, and alignment files return path objects
by default so format-specific tools can process them.

# References

- ENCODE Project Consortium. An integrated encyclopedia of DNA elements in the
    human genome. *Nature*. 2012;489:57-74.
- Kagda MS et al. Data navigation on the ENCODE portal. *Nature
    Communications*. 2025;16:9592.
- [ENCODE REST API](https://www.encodeproject.org/help/rest-api/)
- [ENCODE citation guidance](https://www.encodeproject.org/help/citing-encode/)

# Session information

```{r session-info}
sessionInfo()
```
