---
title: "Query and retrieve ENCODE files"
author: "Zoheb Khan"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Query and retrieve ENCODE files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(encodeUtils)
```

# Overview

`encodeUtils` provides one auditable path from an ENCODE metadata query to
local files, R objects, expression matrices, and a provenance manifest. Search
and file results retain the ENCODE query URL, active filters, retrieval time,
and flattened record metadata.

Install the current development version from GitHub as described in the package
[README](https://github.com/ZohebKhan1/encodeUtils#installation).

# File-selection presets

Presets provide format and output-type preferences for common assays. Inspect a
preset before using it when the exact priorities matter.

```{r presets}
encode_file_presets()
encode_file_presets("rnaseq_gene_quant")
```

# End-to-end RNA-seq workflow

The workflow below finds one released mouse heart RNA-seq experiment, selects
gene quantification files, downloads them to an `encode-data` directory in the
current project, assembles count and TPM matrices, and writes a JSON manifest.

```{r workflow, eval=FALSE}
experiments <- encode_search(
    organism = "mouse",
    assay = "rna-seq",
    organ = "heart",
    limit = 1
)

experiment_table <- encode_results(experiments)
experiment_table[, c(
    "accession", "assay_title", "biosample_summary", "organism", "status"
)]

files <- encode_list_files(experiments)

selected <- encode_select_files(
    files,
    preset = "rnaseq_gene_quant",
    assembly = "mm10",
    replicate_policy = "preferred_processed"
)

selected_table <- encode_results(selected)
selected_table[, c(
    "experiment_accession", "file_accession", "file_format",
    "output_type", "assembly", "file_size_pretty"
)]
selected$excluded
selected$criteria

downloaded <- encode_download(
    selected,
    directory = "encode-data"
)

download_table <- encode_results(downloaded)
download_table[, c(
    "file_accession", "download_status", "local_path",
    "size_verified", "md5_verified"
)]

loaded <- encode_read(
    downloaded,
    values = c("raw_counts", "tpm")
)

loaded$metadata[, c("file_accession", "local_path")]
loaded$row_data
loaded$matrices$raw_counts
loaded$matrices$tpm

manifest <- encode_manifest(
    loaded,
    path = "encode-manifest.json"
)

manifest$retrieval
manifest$attribution[, c(
    "experiment_accession", "file_accession", "file_format", "output_type"
)]
```

## Search and file selection

`encode_search()` accepts common biological fields directly and prints a
compact result. `encode_results()` exposes the complete table for filtering,
joins, or export. A known ENCSR accession can be passed directly to
`encode_list_files()` when metadata discovery is unnecessary.

`encode_select_files()` retains selected rows in `files`, non-selected rows and
reasons in `excluded`, and all applied choices in `criteria`. In the workflow,
`preferred_processed` keeps the highest-priority preset output type within each
experiment without discarding files tied at that priority.

## Transfer and returned objects

`encode_download()` checks individual and total size limits before transfer.
Existing files are not overwritten by default, and replacement files are
installed only after available size and MD5 checks pass. Set `directory = NULL`
to use the persistent package cache.

A completed download table produces one `encode_loaded_files` object whether it
contains one file or several:

- `metadata` contains the complete input file table and provenance.
- `data` contains one native object per file.
- `row_data` describes features aligned to combined matrices.
- `matrices` contains the requested numeric expression matrices.

Combined matrices are created only when every participating table has a
complete, unique feature key. Otherwise, the original tables remain in `data`
without an inferred row-order alignment. Matrix columns use file accessions,
and matrix row names match `row_data`.

## Other supported formats

For a single local path, `encode_read()` returns the native object for that
format. BED, narrowPeak, and broadPeak files return
`GenomicRanges::GRanges` by default; use `as = "data.frame"` for a plain table.
GFF, GTF, BigWig, and BigBed files use `rtracklayer`, and FASTA files use
`Biostrings`, when those packages are installed. FASTQ, BAM, CRAM, and SAM
files remain path objects for dedicated sequence and alignment readers.

Indexed BigWig and BigBed files require a genomic `region` unless
`allow_large = TRUE` is set deliberately.

## Provenance

`encode_manifest()` records the source query, active filters, file metadata,
loaded object classes, ENCODE attribution fields, package version, and R
session. Supplying `path` writes the same structure as JSON.

# Session information

```{r session-info}
sessionInfo()
```
